---
chapter: 1
knit: "bookdown::render_book"
---

# Introduction {#ch-introduction}

<!-- Exploratory data analysis -->
Exploratory Data Analysis (EDA) is the process of the initial summarization and visualization of a dataset. This is a critical first step of checking for realistic values, finding improper data formats, and revealing insights [@tukey_exploratory_1977]. @wickham_r_2017 describes the analyst workflow into discrete steps. First the data must be imported and cleaned which then entered into a iterated cycle of transformations, visualization, and modeling. This work aids the visualization step, focusing specifically on quantitative multivariate data. In implementation of packages also facilitates the transformations and modeling though that is not the primary contributions.

<!-- (ref:ch1fig1-cap) Data analysis workflow [@wickham_r_2017]. This work focuses primarily on visualization of multivariate data. -->
<!-- ```{r ch1fig1, echo=F, out.width="100%", fig.cap = "(ref:ch1fig1-cap)"} -->
<!-- nsSafeIncGraphic("./figures/ch1_fig1_data_analysis_workflow.PNG") -->
<!-- ``` -->

<!-- Dimensionality increases, making data viz hard -->
As modern datasets have grown in complexity, multivariate data has become ubiquitous. Multivariate data is found in physics, biology, social sciences, and manufacturing  [@wang_mapping_2018; @huber_orchestrating_2015; @brown_confirmatory_2015; @evans_multivariate_2017]. As the number of variables in the data increases, however, it becomes increasingly difficult to visualize the relationships between the variables and discern the structure of the data set. This thesis addresses the visualization of multivariate data and the analysis of multivariate data using the above iterated workflow.

<!-- Shadow analogy segue to linear proj. -->
One of the most common and successful approaches to visualize multivariate data is to project multiple dimensions of the data onto two dimensions and then display the projected data. In the same way that 3D object casts a 2D shadow, these projections cast a profile orientation of the data onto a 2D plane.

<!-- Linear projections -->
Our starting point are linear projections. These use a linear combination of variables, called a basis, to define each dimension of the projected space. Each dimension combines multiple variables. This is in contrast to looking at say, each pair of variables in isolation. The basis of linear projections is frequently illustrated with a biplot [@gabriel_biplot_1971]. The biplot shows the magnitude and angle each variable contributes to the resulting display dimensions inscribed in a unit circle such as in Figure \@ref(fig:ch1fig2).

(ref:ch1fig2-cap) Two linear projections of penguins data. Biplot circles depict the basis; the direction and magnitudes of the contributions of the variables. For instance, on the left frame the direction of separation of the orange cluster is mostly in the direction that `bl` contributes meaning that variable is sensitive to the separation of this cluster. The separation of the purple cluster is attributed to `bd` primarily and `fl` and `bm` to a smaller extent. There are also many other linear orientations that do not resolve structures of interest such as cluster separation (right frame). This is illustrated on Palmer Penguin data [@gorman_ecological_2014; @horst_palmerpenguins_2020] where three species of penguins were measured across four physical variables: bill length (`bl`), bill depth (`bd`), flipper length (`fl`), and body mass (`bm`).
```{r ch1fig2, echo=F, out.width="100%", fig.cap = "(ref:ch1fig2-cap)"}
nsSafeIncGraphic("./figures/ch1_fig2_penguin_cl_sep.png")
```

<!-- Talk about features, structure and sensitivity -->
There are many features that an analyst may be interested in when analyzing multivariate data. Shape and spread, identifying clusters, outliers, and irregularity are some of the most common. While linear projections can show these features, not all projections will reveal the feature. Thus the choice of projection is important. Furthermore, the analyst is interested in understanding how the variables contribute to revealing the feature. For instance, Figure \@ref(fig:ch1fig2) shows two linear projections of penguin data. The color and shape of the data points distinguish the three penguin species. The linear projection used in the left frame does not distinguish the three species while that used in the right frame shows a significant separation.

<!-- Tours, animated linear projections -->
It therefore makes sense for an analyst to explore different linear projections. However, it can be difficult to understand the affect of the projection and to keep track of data points if the analyst is shown several projections with no relationship to one another.  The _tour_ is a dynamic visualization that overcomes this difficulty [@cook_grand_2008; @lee_state_2021]. It is a class of linear projections that animate small changes to the projection basis. A key feature of the tour is the visual permanence and trackability of the points through the frames. In the shadow analogy, an object such as a barstool will cast a circular shadow if the light is directly above the seat. However, such a shadow does not give the observer sufficient information as it could arise from any number of shapes that contain profiles of spheres, cylinders, or circles. However, if the stool was rotated, its legs would show in the shadow quickly giving an intuitive interpretation of the object. Similarly, the rotation of a data object yields information about its structure.

<!-- Manual tour -->
A crucial component for _human-in-the-loop_ analysis [@karwowski_international_2006] is the ability for the analyst to manually steer the tour. This allows, for example, the analyst  to explore what happens to the structure when one variable is removed or the contribution of another is increased. @cook_manual_1997 introduced the _manual tour_, offering user control over the basis. By selecting a variable and initializing an additional manipulation dimension to an embedding, the contribution of the variable can be controlled.

<!-- Radial tour -->
The manual tour allowed the analyst to control both the angle and magnitude that a variable contributes to the projection. However, controlling the magnitude is generally more meaningful as angular manipulations effectively rotate the projection, changing the relative position, but not the contribution of a variable. Because of this, this work focuses on a specific manual tour, the _radial tour_, where the angle of the manipulated variable is fixed, but the analyst can vary its magnitude, changing its radius. Figure \@ref(fig:ch1fig3) shows a radial tour on the penguins data varying the contribution of bill length. In frames 1 and 2 there is a large contribution from bill length, and all species are distinguished. In frame 3, however, bill length is removed, and the separation between orange and green clusters has collapsed; we say this variable is sensitive to the separation of these two clusters.

(ref:ch1fig3-cap) A radial tour changing the contribution of bill length (`bl`). When bill length has a considerable contribution, the clusters of orange and green are separated (left and middle). When its contribution is removed, the clusters overlap (right). Because of this, we say that bill length is sensitive to the separation of these two species. An animated version can viewed at [vimeo.com/676723431](https://vimeo.com/676723431).
```{r ch1fig3, echo=F, out.width="100%", fig.cap = "(ref:ch1fig3-cap)"}
nsSafeIncGraphic("./figures/ch1_fig3_penguin_manualtour.png")
```


## Research questions
<!-- Hypothesis statement -->

Discerning variable sensitivity to the structure is a crucial step in understanding a variables contribution to features or differences of interest to the analyst. We conjecture that the user interaction afforded by the radial tour should allow for a more precise exploration of this structure by testing the variable sensitivity to that structure. The over-arching question of interest can, therefore, be stated as:

__Can the radial tour, with user control of the basis, help analysts understand the variable sensitivity of structure in the projection?__

While @cook_manual_1997 sketched the theoretical basis for the manual (and hence radial) tour, some details are missing. Furthermore, we lack a publicly available implementation, fully featured interface design, implementation notes, and have no evaluation of its performance over alternatives.

RQ 1. __How do we define and implement a user interface and interactions for the radial tours to add and remove variables smoothly from a 2D linear projection of data?__

<!-- Most perceptible way to visualize multivariate spaces -->
At present the radial tour is not used by analysts. Instead they would use a single projection to understand the structure, almost always the principal components analysis (PCA) which chooses the basis that shows the most variation. Another approach is to use a _grand tour_. This randomly selects target bases and then generates an animation that interpolates between these target bases. Neither PCA nor the grand tour provide a means for manually manipulating a desired variable's contribution to the basis. We wish to investigate if the ability for the analyst to steer the basis in the radial tour facilitates better understand variable sensitivity to the structure. 

RQ 2. __Does the use of the interactive radial tour improve analysts understanding of the relationship between variables and structure in 2D linear projections compared to existing approaches?__

<!-- Interpretability crisis of the nonlinear models -->
Complex nonlinear models are also being applied more frequently to predict or classify from many predictors. While these models lead to increased accuracy over linear models, they suffer from a loss of the interpretability of their variables. One aspect of eXplainable Artificial Intelligence [XAI, @adadi_peeking_2018; @arrieta_explainable_2020] tries to preserve the interpretability of such models through local explanations. These explanations are essentially linear variable importance in the vicinity of one observation of a model. That is, the extent that variables help the model explain the difference between the observed means and this observation's prediction. The user control from the radial tour potentially allows an analyst to understand better the model and the support of these local explanations.

RQ 3. __Can the radial tours be used in conjunction with local explanations to improve the interpretability of black-box models?__


## Methodology

The research corresponding with RQ 1 entails _algorithm & software design_ [@kleinberg_algorithm_2006] adapting the algorithm from @cook_manual_1997.

To address RQ 2, we use  _experimental design_ [@winer_statistical_1962]. We must define a task and measure suitable to evaluate the radial tour against alternatives. The experimental factors and their levels must be selected and randomly assigned to explore the efficacy of user-controlled radial tours compared with two benchmark methods.

The research responding to RQ 3 involves _design science_ [@hevner_design_2004]. It is not obvious how to combine a radial tour with a nonlinear model. A local explanation approximates the linear variable importance in the vicinity of one observation. We must develop a novel interactive visualization that accommodate two aspects. First, it should visually facilitate the selection of observations to explore. Secondly, extend the biplot to show the distribution of the local explanations and examine the variable sensitivity to the structure identified in local explanation with the radial tour.


## Contributions

The contributions resulting from the research to address these research questions can be split into scientific knowledge and software contributions: 


### Scientific knowledge

- Radial tour algorithm
    - Refined and clarified the steps to producing a radial tour based on the use of the Rodrigues' rotation formula extending the approach from @rodrigues_lois_1840.
    - The algorithm makes embedding the radial tour in animation system as well as interactive system simpler.
    - Provides new examples of usage.
- A user study comparing the radial tour's efficacy against two alternatives--PCA and the grand tour, the first empirical evaluation of the radial tour.
    - Creation of supervised classification task to assess the variable attribution to the separation of two cluster.
    - As tested over experimental factors: location, shape, and dimensionality.
    - Definition of an accuracy measure to evaluate this task.
    - Results: strong evidence that the radial tour increases the accuracy of this task by a sizable amount and minor evidence to suggest a moderate increase in accuracy of the grand tour over Principal Component Analysis (PCA).
    - Mixed model regression helps to attribution the source of the error accounting for the variability of participants skill and the difficulty from random simulation.
- A novel interactive visualization using the radial tour to explore the variable sensitivity of local explanations from nonlinear models, part of XAI.
    - A global view approximates the variable space, attribution space, and model information side-by-side serves to identify a primary observation of interest.
    - This observation's normalized variable attribution is used as a projection basis.
    - Explore the support of the local explanation; using the radial tour, the variable sensitivity to the structure identified test the support of the explanation.


### Software

- __spinifex__, an __R__ package for transforming data, performing radial tours, and the layered composition of any tours.
    - Facilitates the transformation of numeric variable in the data.
    - Identify various bases finding various features of the data.
    - Creation of manual tours allow analyst steering of the basis to explore the variable sensitivity to structure.
    - Layered composition of tour displays that mirrors the approach in __ggplot2__ [@wickham_ggplot2_2016], interoperable with tours made from __tourr__ [@wickham_tourr:_2011].
    - Interactive __shiny__ [@chang_shiny_2021] application to preprocess data and explore. Users can choose from six supplied datasets or upload their own.
    - Vignettes and code examples help users get up to speed.
    - Introduces an interactive application to preprocess data and explore. Users can choose from six supplied datasets or upload their own.
- __cheem__, an __R__ package that facilitates the exploration of local explanations of nonlinear models:
    - Preprocessing; given a tree-based model, calculate the tree SHAP local explanation (via __treeshap__ [@kominsarczyk_treeshap_2021]) of all observations, and find statistics to accent the separability of this space.
    - Visualization of approximations of the data space, attribution space, and model residual information side-by-side with linked brushing, hover tooltips, and tabular display facilitates the selection of observations to explore.
    - Use of the radial tour varies the contribution of variable test the support of the variable contribution in agreement with the explanation.
    - Interactive application facilitates this analysis for several prepared datasets or user preprocessed data.
    - A vignette and code examples help users get up to speed.


## Thesis structure

The remainder of the thesis is organized as follows: Chapter \@ref(ch-background) covers various visualizing techniques before introducing related studies and then nonlinear models and their interpretability issues. Chapter \@ref(ch-spinifex) discusses the theory and implementation of the manual tour in the package __spinifex__. Chapter \@ref(ch-userstudy) discusses a user study evaluating the radial, manual tour's efficacy compared with PCA and the grand tour. Chapter \@ref(ch-cheem) extends the use of radial tours to improve the interpretability of nonlinear models. Lastly, Chapter \@ref(ch-conclusion) concludes with some takeaways and a discussion of possible extensions.
