---
chapter: 2
knit: "bookdown::render_book"
---

# Background {#ch-background}

This section starts with setting the scope for the type of data we are focusing on, gives a brief motivation then works through previous multivariate visualizations and their scalability. By the end, the focus narrows to linear projections, and in particular, the class of animated linear projections known as the tour.

<!-- note on data and context -->
For our purposes I will be focusing on the case where data $X_{nxp}$ contains $n$ observations of $p$ variables, is complete with no missing values, variables are numeric (ideally not ordinal levels), and $n>p$ typically many more observations than variables. While I write as though always operating on the original variable space, these methods could similarly be applied to component spaces or feature decompositions of data not fitting this format. In the case of the linear projection let, $Y_{nxd} = X_{nxp} \cdot A_{pxd}$ be the embedding of the data mapped by the basis $A$, where $d<p$. When $p$ is large, say over 10 or 20 variables, the viewing space is quite large. In these cases, a PCA initialization step is commonly used where the variables are approximated as fewer principal components, and this reduced space can be viewed, albeit with the disadvantage of having another linear mapping back to the original space. @wickham_visualizing_2015 also view model spaces and features while urging to prefer visualizing data-space directly.

<!-- Better than numerical summarization alone -->
Visualization is much more robust than numerical summarization alone [@anscombe_graphs_1973; @matejka_same_2017]. In these studies, data sets have the same summary statistics yet contain obvious visual trends and shapes that could go completely unheeded if plotting is foregone. Data visualization is fundamental to EDA and quickly evaluating data supports ensuring that models are suitable.

(ref:ch1fig2-cap) Starting from the profile of a dinosaur, observations are allowed to drift (by iterated simulated annealing) toward 12 patterns provided that they stay close to the original statistics [@matejka_same_2017]. Visualization of data yields stark designs that are easy to miss in numerical summarization.

```{r ch1fig2, echo=F, out.width='70%', fig.cap = "(ref:ch1fig2-cap)"}
nsSafeIncGraphic("./figures/ch1_fig2_matejka17fig.PNG")
```

## Scatterplot matrices

<!-- Penguins data -->
The work in @grinstein_high-dimensional_2002 gives a good taxonomy of high-dimensional visualization. We will follow a few examples conclude with tour methods. Broadly speaking, we are concerned with the question "How can an analyst visualize arbitrary $p-$ dimensions?". To illustrate some of the options for data I use the penguins data [@gorman_ecological_2014; @horst_palmerpenguins_2020]. It contains 333 observations of four physical measurements across three species of penguins observed near Palm Station, Antarctica.

<!-- SPLOM -->
Viewing as many univariate histograms or density curves is one method. Similarly, one could look at all variable pairing as scatter plots. This forms the crux of the scatterplot matrices, also known as SPLOM [@chambers_graphical_1983]. In a scatterplot matrix, variables are displayed across the columns and rows. The diagonal elements show the univariate densities while off-diagonal positions show scatterplot pairs. This is useful for getting a handle on the support of the variables but is not going to scale well with dimension and is not a suitable audience-ready display. It is very busy and doesn't draw attention to any one spot. @munzner_visualization_2014 reminds us to abstract all of the cognitive work out of the visual, allowing the audience to focus on seeing the evidence supporting the claim.

(ref:penguinsplom-cap) Scatterplot matrix of penguins data. This is a good initial step, but will not scale well at $p$ increases.

```{r penguinsplom, echo=F, out.width='100%', fig.cap = "(ref:penguinsplom-cap)"}
nsSafeIncGraphic("./figures/ch2_fig1_penguin_splom.pdf")
```

## Parallel coordinate plots

<!-- PCP & observation based visuals -->
Alternatively, we could consider a class of observation-based visuals. In parallel coordinate plots [@ocagne_coordonnees_1885], variables are arranged horizontally and lines connect observations with the height mapping to the quantile or z-value for each variable.  This scales much better with dimensions but poorly with observations. It also suffers from an asymmetry with the variable order. That is, changing the order of the variable will lead people to very different conclusions. The x-axis is also used to display variables rather than the values of the observations. This restricts the amount of information that can be interpreted between variables. Munzner asserts that position is the more human-perceptible channel for encoding information; we should like to reserve it for the values of the observations. The same issues persist across other observation-based displays such as radial variants, pixel-oriented visuals, and Chernoff faces [@keim_designing_2000; @chernoff_use_1973]. These visuals are better suited for the $n<p$ case with more variables than observations.

(ref:penguinpcp-cap) Parallel coordinate plots of penguins data. This does not scale well with observations, suffers from asymmetry with the variable ordering, and horizontal position is used for variables rather than observation levels.

```{r penguinpcp, echo=F, out.width='100%', fig.cap = "(ref:penguinpcp-cap)"}
nsSafeIncGraphic("./figures/ch2_fig2_penguin_pcp.pdf")
```


## Dimension reduction

Ultimately, we will need to turn to dimension reduction to create a compelling visual allowing audiences to focus on features with contributions from multiple variables. Dimension reduction is separated into two categories, linear and non-linear. The linear case spans all affine mathematical transformations, essentially any mapping where parallel lines stay parallel. Non-linear transformations are the complement of the linear case, think transformations containing exponents or interacting terms. Examples in low dimensions are relatable. For instance, shadows are examples of linear projections where a 3-dimensional object casts a 2D projection, the shadow. Our vision at any one instance of time, or a picture, is also a 2D projection. An example of a non-linear transformation is that of 2D representation of the globe. There are many different ways (and features to optimize) to distort the surface to display as a map. The most common may be rectangular displays where the area is distorted more and more with the distance away from the equator. Other distortions are created when the surface is unwrapped as a long ellipse. Yet others create non-continuous gaps in oceans to minimize the distortion of countries.

Non-linear techniques often have hyperparameters that affect how the spaces are distorted to fit into fewer dimensions. To quote Anastasios Panagiotelis, "All non-linear projections are wrong, but some are useful", a play on George Box's quote about models. Non-linear techniques distort the space in unclear ways, and what is more, they can introduce features not in the data depending on the selection of hyperparameters. The presence of structure in a non-linear model is necessary but not sufficient to conclude the existence of a structure in the data.

Unfortunately, there is no free lunch here. An increase in the original data dimensions will lead to a $p-d$-dimensional viewing space in the linear case and an increasingly perturbed and distorted space in non-linear techniques. The intrinsic dimensionality of data is the number of variables needed to minimally represent the data [@grinstein_high-dimensional_2002]. Intrinsic data dimensionality is an important aspect of dimension reduction that does not have to end in visual, but is also a common part of factor analysis and preprocessing data. Consider a Psychology survey consisting of 100 questions about the Big Five personality traits. The data consists of 100 variables, while the theory would suggest the intrinsic dimensionality is five. The data likely picks up on other aspects and may be better summarized in with eight or ten dimensions. If this were the case, reducing the data to this space would be necessary to gate the exponentially increasing view time.


## Tours, animated linear projections

A single linear projection is a resulting space from the data multiplied by the basis. The basis is an orthonormal matrix (the orientation of the unit origin) mapping the data space to a lower dimension. A data visualization tour animates many such projections through small changes in the basis. In the bar stool shadow analogy, structural information about a hidden object was gained by watching its shadow change due to its rotation. An analyst similarly gains information about the data object by watching continuous changes to the basis (the rotation of the data). Originally in a _grand_ tour [@asimov_grand_1985] several target frames are randomly selected and then interpolated along their geodesic path.

(ref:ch1fig3-cap) Illustration of the grand tour selecting random target frames (grey) connected via geodesically interpolated frame (white). Figure from @buja_computational_2005.

```{r ch1fig3, echo=F, out.width='100%', fig.cap = "(ref:ch1fig3-cap)"}
nsSafeIncGraphic("./figures/ch1_fig3_buja05fig.PNG")
```

There are various types of tours which are classified by the generation of their basis paths. A _guided_ tour uses simulated annealing to move progressively closer to an objective function in the embedded space [@hurley_analyzing_1990]. A more comprehensive discussion and review of tours can be found in the works of @cook_grand_2008 and @lee_state_2021.

Tours are used for a couple of salient features: maintains transparency back to the original variable space, and the persistence data points from frame to frame convey more information than looking at discrete jumps to other bases. Because of these features, tours are a good method to extend the visualization of data space as dimensionality increases.

The work below covers manual tours [@cook_manual_1997; @spyrison_spinifex_2020] in Chapter \@ref(ch-spinifex). Radial tours are one type of manual tour where the contribution of one variable is extended radially to a full contribution, removed completely, then restored to its original contribution. Chapter \@ref(ch-userstudy) compares the efficacy of the radial tour as compared with PCA and the grand tour in a user study. Lastly, Chapter \@ref(ch-cheem) extends the use of the radial tour to evaluate the local explanation of black-box models.



