---
chapter: 3
knit: "bookdown::render_book"
---
```{r, include=FALSE}
## Original R journal paper:
if(F)
  file.edit("../spinifex_paper/paper/spyrison-cook.Rmd")
## Creating figures in script (not ran upon compilation)
if(F)
  file.edit("./figures/_ch3_spinifex.r")
```

# spinifex: an R package for creating user-controlled animated linear projections {#ch-spinifex}

<!-- Segue and disclose change -->
This chapter introduces manual tours that allows analysts to influence the contributions to a projection. This feature is absent from previous linear embeddings. This work has changed from its publication, primarily to incorporate and discuss the subsequent 'ggproto' API that greatly extends he flexibility of the geometric displays of tours.

<!-- Abstract -->
Dynamic low-dimensional linear projections of multivariate data known as _tour_ provide an important tool for exploring multivariate data and models. The R package __tourr__ provides functions for several types of tours: grand, guided, little, local, and frozen. Each of these can be viewed in a development environment, or their basis array can be saved for later consumption. This paper describes a new package, __spinifex__, which provides a manual tour of multivariate data. In a manual tour an analyst controls the contribution of a variable to the projection. Controlled manipulation is important to explore a variables sensitivity of structure in an identified feature. The use of the manual tour is applied to particle physics data to illustrate the sensitivity of structure in a projection to specific variable contributions. Additionally, we create a 'ggproto' API for composing any tour that mirrors the layered additive approach of __ggplot2__. Tours can then be animated and exported to a variety of formats with __plotly__ or __gganimate__.


## Introduction

<!-- Tours -->
Exploring multivariate spaces is a challenging task, increasingly so as dimensionality increases. Traditionally, static low-dimensional projections are used to display multivariate data in two dimensions including principal component analysis, linear discriminant spaces or projection pursuit. These are useful for finding relationships between multiple variables, but they are limited because they show only a glimpse of the high-dimensional space. An alternative approach is to use a tour [@asimov_grand_1985] of dynamic linear projections to look at many different low-dimensional projections. Tours can be considered to extend the dimensionality of visualization, which is important as data and models exist in more than 3D. The package __tourr__ [@wickham_tourr:_2011] provides a platform for generating tours. It can produce a variety of tours, each paired with a variety of possible displays. A user can make a grand, guided, little, local or frozen tour, and display the resulting projected data as a scatterplot, density plot, histogram, or even as Chernoff faces if the projection dimension is more than 3.

<!-- Manual tour -->
This work adds a manual tour to the collection. The manual tour was described in @cook_manual_1997 and allows a user to control the projection coefficients of a selected variable in a 2D projection. The manipulation of these coefficients allows the analyst to explore their sensitivity to the structure within the projection. As manual tours operate on only one variable at a time, they are particularly useful once a feature of interest has been identified. 

<!-- interesting features and guided tours -->
One way to identify "interesting" features is with the use of a guided tour [@cook_grand_1995]. Guided tours select a very specific path, which approaches a projection that optimizes an objective function. The optimization used to guide the tour is simulated annealing [@kirkpatrick_optimization_1983]. The direct optimization of a function allows guided tours to rapidly identify interesting projection features given the relatively large parameter-space. After a projection of interest is identified, an analyst can then use the "finer brush" of the manual tour to control the contributions of individual variables to explore the sensitivity they have on the structure visible in the projection.

<!-- Paper outline -->
The paper is organized as follows. Section \@ref(sec:algorithm) describes the algorithm used to perform a radial manual tour as implemented in the package __spinifex__. Section \@ref(sec:display) explains how to generate an animation of the manual tour using the animation frameworks offered by __plotly__ [@sievert_interactive_2020] and __gganimate__ [@pedersen_gganimate_2020]. Package functionality and code usage following the order applied in the algorithm follows in section \@ref(sec:usage). Section \@ref(sec:usecases) illustrates how this can be used for sensitivity analysis applied to multivariate data collected on high energy physics experiments [@wang_mapping_2018]. Section \@ref(sec:discussion) summarizes this paper and discusses potential future directions.


<!-- Algorithm outline -->
## Algorithm {#sec:algorithm}

The algorithm to conduct a manual tour interactively, by recording mouse/cursor motion, is described in detail in @cook_manual_1997. Movement can be in any direction and magnitude, but it can also be constrained in several ways:

- *radial*: fix the direction of contribution, and allow the magnitude to change.
- *angular*: fix the magnitude, and allow the angle or direction of the contribution to vary.
- *horizontal*, *vertical*: allow rotation only around the horizontal or vertical axis of the current 2D projection.

The algorithm described here produces a **radial** tour as an *animation sequence*. It takes the current contribution of the chosen variable, and using rotation brings this variable fully into the projection, completely removes it, and then returns to the original position.


### Notation

The notation used to describe the algorithm for a 2D radial manual tour is as follows:

- $\textbf{X}$, the data, an $n \times p$ numeric matrix to be projected to 2D.
- $\textbf{B} = (B_1,~ B_2)$, any 2D orthonormal projection basis, $p \times 2$ matrix, describing the projection from $\mathbb{R}^p \Rightarrow \mathbb{R}^2$. This is called this the "original projection" because it is the starting point for the manual tour.
- $k$, is the index of the manipulation variable or manip var for short.

- $\textbf{e}$, a 1D basis vector of length $p$, with 1 in the $k$-th position and 0 elsewhere.
- $\textbf{M}$ is a $p \times 3$ matrix, defining the 3D subspace where data rotation occurs and is called the manip(ulation) space.
- $\textbf{R}$, the 3D rotation matrix, for performing unconstrained 3D rotations within the manip space, $\textbf{M}$.
- $\theta$, the angle of in-projection rotation, for example, on the reference axes; $c_\theta, s_\theta$ are its cosine and sine.
- $\phi$, the angle of out-of-projection rotation, into the manip space; $c_\phi, s_\phi$ are its cosine and sine. The initial value for animation purposes is $\phi_1$.
- $\textbf{U}$, the axis of rotation for out-of-projection rotation orthogonal to $\textbf{e}$.
- $\textbf{Y}$, the resulting projection of the data through the manip space, $\textbf{M}$, and rotation matrix, $\textbf{R}$.

<!-- operate on bases -->
The algorithm operates entirely on projection bases and incorporates the data only when making the projected data plots, in light of efficiency.


### Steps

#### Step 0) Setup

<!-- describe data. -->
The flea data (@lubischew_use_1962), available in the __tourr__ package, is used to illustrate the algorithm. The data contains 74 observations on 6 variables, which are physical measurements made on flea beetles. Each observation belongs to one of three species.

<!-- Projection basis -->
An initial 2D projection basis must be provided. A suggested way to start is to identify an interesting projection using a projection pursuit guided tour. Here the holes index is used to find a 2D projection of the flea data, which shows three separated species groups. Figure \@ref(fig:ch3fig1) shows the initial projection of the data. The left panel displays the projection basis ($\textbf{B}$) and can be used as a visual guide of the magnitude and direction that each variable contributes to the projection. The right panel shows the projected data, $\textbf{Y}_{[n,~2]} ~=~ \textbf{X}_{[n,~p]} \textbf{B}_{[p,~2]}$. The color and shape of points are mapped to the flea species. This plot is made using the `view\_basis()` function in __spinifex__, which generates a __ggplot2__ [@wickham_ggplot2_2016] object. 

```{r ch3fig1, fig.cap = "Biplot of the initial 2D projection: representation of the basis (left) and resulting data projection (right) of standardized flea data. The color and shape of data points are mapped to the species of flea beetle. The basis was produced by a projection pursuit guided tour, with the holes index. The contribution of the variables aede2 and tars1 approximately contrasts the other variables. The visible structure in the projection are the three clusters corresponding to the three species."}
nsSafeIncGraphic("./figures/ch3_fig1_biplot.pdf")
```


#### Step 1) Choose manip variable

<!-- select a manip var-->
In figure \@ref(fig:ch3fig1) the contribution of the variables tars1 and aede2 mostly contrast the contribution of the other four variables. These two variables combined contribute in the direction of the projection where the purple cluster is separated from the other two clusters. The variable aede2 is selected as the manip var, the variable to be controlled in the tour. The question that will be explored is: how important is this variable to the separation of the clusters in this projection?


#### Step 2) Create the 3D manip space

<!-- Zero Vect, manip sp -->
Initialize the coordinate basis vector as a zero vector, $\textbf{e}$, of length $p$, and set the $k$-th element to 1. In the example data, aede2 is the fifth variable in the data, so $k=5$, set $e_5=1$. Use a Gram-Schmidt process to orthonormalize the coordinate basis vector on the original 2D projection to describe a 3D manip space, $\textbf{M}$.

\begin{align*}
  e_k &\leftarrow 1 \\ 
  \textbf{e}^*_{[p,~1]} &= \textbf{e} - \langle \textbf{e}, \textbf{B}_1 \rangle \textbf{B}_1 - \langle \textbf{e}, \textbf{B}_2 \rangle \textbf{B}_2 \\ 
  \textbf{M}_{[p,~3]} &= (\textbf{B}_1,\textbf{B}_2,\textbf{e}^*)
\end{align*}

<!-- What the manip space provides -->
The manip space provides a 3D projection from $p$-dimensional space, where the coefficient of the manip var can range completely between [0, 1]. This 3D space serves as the medium to rotate the projection basis relative to the selected manipulation variable. Figure \@ref(fig:ch3fig2) illustrates this 3D manip space with the manip var highlighted. This representation is produced by calling the `view\_manip\_space()` function. This diagram is purely used to help explain the algorithm.

```{r ch3fig2, fig.cap = "Illustration of a 3D manip space, the projection plane is shown as blue circle extending into and out of the display. A manipulation direction is initalized, the red circle, orthogonal to the projection plane. This allows the selected variable, aede2, a means to change it's contribution back to the projection plane. The other variables contributions rotate into this space as well, preserving the orthagonal structure, but are omitted in the manipulation dimension for simplicity."}
nsSafeIncGraphic("./figures/ch3_fig2_manip_sp.pdf")
```


#### Step 3) Defining a 3D rotation

<!-- illustration of axis manip -->
The basis vector corresponding to the manip var (red line in Figure \@ref(fig:ch3fig2)), can be operated like a lever anchored to the origin. This is the process of the manual control, that rotates the manip variable into and out of the 2D projection (Figure \@ref(fig:ch3fig3)). As the variable contribution is controlled, the manip space rotates, and the projection onto the horizontal projection plane correspondingly changes. This is a manual tour. Generating a sequence of values for the rotation angles produces a path for the rotation of the manip space.

<!-- describe manip var path -->
For a radial tour, fix $\theta$, the angle describing rotation within the projection plane, and compute a sequence for $\phi$, defining movement out of the plane. This will change $\phi$ from the initial value, $\phi_1$, the angle between $\textbf{e}$ and its shadow in $\textbf{B}$, to a maximum of $0$ (manip var fully in projection), then to a minimum of $\pi/2$ (manip var out of projection), before returning to $\phi_1$.

<!-- define the rotation matrix -->
Rotations in 3D can be defined by the axes they pivot on. Rotation within the projection, $\theta$, is rotation around the $Z$ axis. Out-of-projection rotation, $\phi$, is the rotation around an axis on the $XY$ plane, $\textbf{U}$, orthogonal to $\textbf{e}$. Given these axes, the rotation matrix, $\textbf{R}$ can be written as follows, using Rodrigues' rotation formula (originally published in @rodrigues_lois_1840):

  \begin{align*}
    \textbf{R}_{[3,~3]} 
    &= \textbf{I}_3 + s_\phi\*\textbf{U} + (1-c_\phi)\*\textbf{U}^2 \\
        &=
    \begin{bmatrix}
      1 & 0 & 0 \\ 
      0 & 1 & 0 \\ 
      0 & 0 & 1 \\
    \end{bmatrix} +
    \begin{bmatrix}
      0 & 0 & c_\theta s_\phi \\
      0 & 0 & s_\theta s_\phi \\
      -c_\theta s_\phi & -s_\theta s_\phi & 0 \\
    \end{bmatrix} +
    \begin{bmatrix}
      -c_\theta (1-c_\phi) & s^2_\theta (1-c_\phi) & 0 \\
      -c_\theta s_\theta (1-c_\phi) & -s^2_\theta (1-c_\phi) & 0 \\
      0 & 0 & c_\phi-1 \\
    \end{bmatrix} \\
    &= 
    \begin{bmatrix}
      c_\theta^2 c_\phi + s_\theta^2 &
      -c_\theta s_\theta (1 - c_\phi) &
      -c_\theta s_\phi \\
      -c_\theta s_\theta (1 - c_\phi) &
      s_\theta^2 c_\phi + c_\theta^2 &
      -s_\theta s_\phi \\
      c_\theta s_\phi &
      s_\theta s_\phi &
      c_\phi
    \end{bmatrix} \\
\end{align*}

\noindent where

\begin{align*}
  \textbf{U} &= (u_x, u_y, u_z) =
  (s_\theta, -c_\theta, 0) \\ 
  &=
  \begin{bmatrix}
  0 & -u_z & u_y  \\
  u_z & 0 & -u_x \\
  -u_y & u_x & 0 \\
  \end{bmatrix} =
  \begin{bmatrix}
    0 & 0 & -c_\theta \\
    0 & 0 & -s_\theta \\
    c_\theta & s_\theta & 0 \\
  \end{bmatrix} \\
  \end{align*}
  
<!-- The term $(1-c_\phi)$ is used by convention, but $2sin^2(\phi/2)$ is more computationally robust. -->


#### Step 4) Creating an animation of the radial rotation

<!--  phi transform and animation -->
The steps outlined above can be used to create any arbitrary rotation in the manip space. To use these for sensitivity analysis, the radial rotation is built into an animation where the manip var is rotated fully into the projection, completely out, and then back to the initial value. This involves allowing $\phi$ to vary between $0$ and $\pi/2$, call the steps $\phi_i$. 


```{r ch3fig3, fig.cap = "Select frames highlight the animation of a radial manual tour manipulating aede2: (1) original projection, (2) full contribution, (3) zero contribution, before returning to the original original contribution."}
nsSafeIncGraphic("./figures/ch3_fig3_filmstrip.pdf")
```


<!-- Sequence of $\phi_i$ -->
1. Set initial value of $\phi_1$ and $\theta$: $\phi_1 = \cos^{-1}{\sqrt{B_{k1}^2+B_{k2}^2}}$, $\theta = \tan^{-1}\frac{B_{k2}}{B_{k1}}$. Where $\phi_1$ is the angle between $\textbf{e}$ and its shadow in $\textbf{B}$.
2. Set an angle increment ($\Delta_\phi$) that sets the step size for the animation, to rotate the manip var into and out of the projection. (Note: Using angle increment, rather than a number of steps, to control the movement, is consistent with the tour algorithm as implemented in the __tourr__).
3. Step towards $0$, where the manip var is completely in the projection plane.
4. Step towards $\pi/2$, where the manip variable has no contribution to the projection.
5. Step back to $\phi_1$.

In each of the steps 3-5, a small step may be added to ensure that the endpoints of $\phi$ ($0$, $\pi/2$, $\phi_1$) are reached. 

#### Step 5) Projecting the data {#sec:display}

<!-- the reminder of basis operation and now apply data -->
The operation of a manual tour is defined on the projection bases. Only when the data plot needs to be made is the data projected into the relevant basis. 

\begin{align*}
  \textbf{Y}^{(i)}_{[n,~3]} &= \textbf{X}_{[n,~p]} \textbf{M}_{[p,~3]} \textbf{R}^{(i)}_{[3,3]}
\end{align*}

<!-- plot XY in seq for animation -->
\noindent where $\textbf{R}^{(i)}_{[3,3]}$ is the incremental rotation matrix, using $\phi_i$. To make the data plot, use the first two columns of \textbf{Y}. Show the projected data for each frame in sequence to form an animation. 

<!-- <!-- Animation on flea data -->
<!-- Figure \@ref(<FIG_REMOVED>)illustrates a manual tour sequence having 15 steps. The projection axes are displayed on the top half, which corresponds to the projected data in the bottom half. When aede2 is removed from the projection, the purple cluster overlaps with the green cluster. This suggests that aede2 is important for distinguishing between these species.  -->

Tours are typically viewed as an animation. The animation of this tour can be viewed online on [github](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/flea_radialtour_mvar5.gif). The page may take a moment to load. Animations can be produced using the function `play\_manual\_tour()`.

```{r FigNotAppearingInThisProj,  fig.cap = "Radial manual tour manipulating aede2 of standardized flea data. The axis for aede2 increases in contribution to the projection, from its initial value to 1, decreasing to 0 and then returning to the initial value. This effects the separation between the purple and green clusters. This shows that aede2 is important for distinguishing the purple species, because the separation disappears when aede2 is not contributing to the projection."}
## There was a figure with 15 frame of the tour, I think leave it out for here, didn't come through very well there.
```

## Package structure and functionality

This section describes the functions available in the package, and how to use them.


### Installation

The __spinifex__ is available from CRAN, and can be installed by: 
```{r installCODEEXAMPLE, eval=F, echo=T}
# Setup:
install.package("spinifex") ## Install from CRAN
library("spinifex") ## Read into session

# Getting started:
## Shiny app for visualizing basic application
run_app("intro")
## View the code vignette
vignette("spinifex_vignette")
```


### Functions

<!-- in terms of workflow -->
Table \@ref(tab:functionsTable) lists the primary functions and their purpose. These are grouped into four types: construction for building a tour path, render to make the plot objects, animation for running the animation, and specialty for providing illustrations used in the algorithm description.

```{r functionsTable, echo=T}
library(kableExtra)

funcs_tib <- tibble::tribble(
  ~Family,      ~Function,                 ~`Related to`,               ~Description,
  "processing", "scale_01/sd",             "-",                        "Scale each column to [0,1]/std dev away from the mean",
  "processing", "basis_pca/olda/odp/onpp", "Rdimtools::do.*",          "Basis of orthogonal component spaces",
  "processing", "basis_half_circle",       "-",                        "Basis with uniform contribution across half of a circle",
  "processing", "basis_guided",            "tourr::guided_tour",       "Silently return the basis from a guided tour",
  "tour path",  "manual_tour",             "-",                        "Basis and interpolation information for a manual tour",
  "tour path",  "save_history",            "tourr::save_history",      "Silent, extended wrapper returning other tour arrays",
  "ggtour",     "ggtour",                  "ggplot2::ggplot",          "Canvas and initalization for a tour animation",
  "ggtour",     "proto_point/text",        "ggplot2::geom_point/text", "Adds observation points/text",
  "ggtour",     "proto_density/2d",        "ggplot2::geom_density/2d", "Adds density curve/2d contours",
  "ggtour",     "proto_hex",               "ggplot2::geom_hex",        "Adds hexigonal heatmap of observations",
  "ggtour",     "proto_basis/1d",          "-",                        "Adds adding basis reference frame in a unit-circle/-rectangle",
  "ggtour",     "proto_origin/1d",         "-",                        "Adds adding 2d basis reference frame",
  "ggtour",     "proto_default/1d",        "-",                        "Convience wrapper for proto_* point/density + basis/1d + origin",
  "ggtour",     "facet_wrap_tour",         "ggplot2::facet_wrap",      "Facets on the levels of variable",
  "ggtour",     "append_fixed_y",          "-",                        "Add/overwrite a fixed vertical position of the animatation",
  "Animating",  "animate_plotly",           "plotly::ggplotly",        "Render as an interactive hmtl widget",
  "Animating",  "animate_gganimate",        "gganimate::animate",      "Render as a .gif, .mp4, or other video format",
  "Animating",  "filmstrip",                "-",                       "Static gpplot faceting on the frames of the aimation"
)
  
knitr::kable(funcs_tib, "latex", caption = "Summary of primary functions.", 
             booktabs = TRUE, linesep = "")
```


### Usage {#sec:usage}

Using the penguins data , available in the package, to illustrate a manual tour, we will illustrate generating a manual tour to explore the sensitivity of a variable separating two clusters.

The function `oblique_frame` can be used to show a projection of the basis, or with the data overlaid. For example, the plots in Figures \@ref(fig:ch3fig1) and \@ref(fig:ch3fig3) were made with code similar to this:

```{r eval=F, echo=T}
## Process peguins data
dat     <- penguins[1:4]
clas    <- penguins$species
bas     <- basis_olda(data = dat, class = clas)
mt_path <- manual_tour(basis = bas, manip_var = 1, data = dat)

## Proto API, creating a ggplot2 plot with frames
ggt <- ggtour(mt_path, angle = .15) +
  proto_point(aes_args = list(color = clas, shape = clas),
              identity_args = list(alpa = .8, size = 1.5)) +
  proto_basis("left") +
  proto_origin()

## Choose an animation type
animate_plotly(ggt, fps = 10)
```

## Use cases {#sec:usecases}

<!-- Introduction of data and original paper -->
@wang_mapping_2018 introduces a new tool, PDFSense, to visualize the sensitivity of hadronic experiments to nucleon structure. The parameter-space of these experiments lies in 56 dimensions, $\delta \in \mathbb{R}^{56}$, and are visualized as 3D subspaces of the 10 first principal components in linear (PCA) and non-linear (t-SNE) embeddings. 

<!-- grand tours on the same data -->
@cook_dynamical_2018 illustrates how to learn more about the structures using a grand tour. Tours can better resolve the shape of clusters, intra-cluster detail, and better outlier detection than PDFSense & TFEP (TensorFlow embedded projections) or traditional static embeddings. This example builds from here, illustrating how the manual tour can be used to examine the sensitivity of structure in a projection to different parameters. The specific 2D projections passed to the manual tour were provided in their work. 

<!-- Data structure -->
The data has a hierarchical structure with top-level clusters; DIS, VBP, and jet. Each cluster is a particular class of experiments, each with many experimental datasets which,  each have many observations of their own. In consideration of data density, we conduct manual tours on subsets of the DIS and jet clusters. This explores the sensitivity of the structure to each of the variables in turn and we present the subjectively best and worst variable to manipulate for identifying dimensionality of the clusters and describing the span of the clusters.


### Jet cluster

<!-- jet cluster, explain dimensionality -->
The jet cluster resides in a smaller dimensionality than the full set of experiments with four principal components explaining 95% of the variation in the cluster [@cook_dynamical_2018]. The data within this 4D embedding is further subsetted, to ATLAS7old and ATLAS7new, to focus on two groups that occupy different parts of the subspace. Radial manual tours controlling contributions from PC4 and PC3 are shown in Figures \@ref(fig:ch3fig4) and \@ref(fig:ch3fig5), respectively. The difference in shape can be interpreted as the experiments probing different phase-spaces. Back-transforming the principal components to the original variables can be done for a more detailed interpretation.

<!-- discussion of findings and which is more insightful -->
When PC4 is removed from the projection (Figure \@ref(fig:ch3fig4)) the difference between the two groups is removed, indicating that it is important for distinguishing experiments. However, removing PC3 from the projection (Figure \@ref(fig:ch3fig5)) does not affect the structure, indicating it is not important for distinguishing experiments. Animations for the remaining PCs can be viewed at the following links: [PC1](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/jetcluster_manualtour_pc1.gif), [PC2](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/jetcluster_manualtour_pc2.gif), [PC3](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/jetcluster_manualtour_pc3.gif), and [PC4](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/jetcluster_manualtour_pc4.gif). It can be seen that only PC4 is important for viewing the difference in these two experiments.

<!-- JetClusterGood -->
```{r ch3fig4, fig.cap="Select frames from a radial tour of PC4 within the jet cluster, with color indicating experiment type: ATLAS7new (green) and ATLAS7old (orange). When PC4 is removed from the projection (frame 10) there is little difference between the clusters, suggesting that PC4 is important for distinguishing the experiments."}
nsSafeIncGraphic("./figures/ch3_fig4_jet_better_pc4.pdf")
```
<!-- JetClusterBad -->
```{r ch3fig5, fig.cap = "Frames from the radial tour manipulating PC3 within the jet cluster, with color indicating experiment type: ATLAS7new (green) and ATLAS7old (orange).  When the contribution from PC3 is changed there is little change in the separation of the clusters, suggesting that PC3 is not important for distinguishing the experiments."}
nsSafeIncGraphic("./figures/ch3_fig5_jet_worse_pc3.pdf")
```


### DIS cluster

<!-- introduce DIS cluster -->
Following @cook_dynamical_2018, to explore the DIS cluster, PCA is recomputed and the first six principal components, explaining 48% of the full sample variation, are used. The contributions of PC6 and PC2 are explored in Figures \@ref(fig:ch3fig6) and \@ref(fig:ch3fig7), respectively. Three experiments are examined: DIS HERA1+2 (green), dimuon SIDIS (purple), and charm SIDIS (orange).

<!-- comparison of the DIS cluster -->
Both PC2 and PC6 contribute to the projection similarly. When PC6 is rotated into the projection, variation in the DIS HERA1+2 is greatly reduced. When PC2 is removed from the projection, dimuon SIDIS becomes more clearly distinct. Even though both variables contribute similarly to the original projection, their contributions have quite different effects on the structure of each cluster, and the distinction between clusters. Animations of all of the principal components can be viewed from the links: [PC1](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/discluster_manualtour_pc1.gif), [PC2](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/discluster_manualtour_pc2.gif), [PC3](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/discluster_manualtour_pc3.gif), [PC4](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/discluster_manualtour_pc4.gif), [PC5](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/discluster_manualtour_pc5.gif), and [PC6](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/discluster_manualtour_pc6.gif).

<!-- DISclusterGood -->
```{r ch3fig6, fig.cap = "Select frames from a radial tour exploring the sensitivity that PC6 has on the structure of the DIS cluster, with color indicating experiment type: DIS HERA1+2 (green), dimuon SIDIS (purple), and charm SIDIS (orange). DIS HERA1+2 is distributed in a cross-shaped plane, charm SIDIS occupies the center of this cross, and dimuon SIDIS is a linear cluster crossing DIS HERA1+2. As the contribution of PC6 is increased, DIS HERA1+2 becomes almost singular in one direction (frame 5), indicating that this cluster has very little variability in the direction of PC6."}
nsSafeIncGraphic("./figures/ch3_fig6_DIS_better_pc6.pdf")
```

<!-- DISclusterBad -->
```{r ch3fig7, fig.cap = "Frames from the radial tour exploring the sensitivity PC2 to the structure of the DIS cluster, with color indicating experiment type: DIS HERA1+2 (green), dimuon SIDIS (purple), and charm SIDIS (orange). As contribution from PC2 is decreased, dimuon SIDIS becomes more distinguishable from the other two clusters, indicating that in its absence PC2 is important for separating this cluster from the others."}
nsSafeIncGraphic("./figures/ch3_fig7_DIS_worse_pc2.pdf")
```

## Discussion {#sec:discussion}

<!-- summary of spinifex  -->
Dynamic linear projections of numeric multivariate data, tours, play an important role in data visualization; they extend the dimensionality of visuals to peek into high-dimensional data and parameter spaces. This research has taken the manual tour algorithm, specifically the radial rotation, used in GGobi [@swayne_ggobi:_2003] to interactively rotate a variable into or out of a 2D projection, and modified it to create an animation that performs the same task. It is most useful for examining the importance of variables, and how the structure in the projection is sensitive or not to specific variables. This functionality available in package __spinifex__. The work complements the methods available in the __tourr__ package.

<!-- summary of application  -->
This work was motivated by problems in physics, and thus the usage was illustrated on data comparing experiments of hadronic collisions, to explore the sensitivity of cluster structure to different principal components. These tools can be applied quite broadly to many multivariate data analysis problems.

<!-- Constraints -->
The manual tour is constrained in the sense that the effect of one variable is dependent on the contributions of other variables in the manip space. However, this can be useful to simplify a projection by removing variables without affecting the visible structure. Defining a manual rotation in high dimensions is possible using Givens rotations and Householder reflections as outlined in @buja_computational_2005. This would provide more flexible manual rotation, but more difficult for a user because they have the choice (too much choice) of which directions to move. 

<!-- 3D and function vis -->
Another future research topic could be to extend the algorithm for use on 3D projections. With the current popularity and availability of 3D virtual displays, this may benefit the detection and understanding of the higher dimensional structure, or enable the examination of functions.

<!-- dynamic interaction/gui -->
Having a graphical user interface would be useful for making it easier and more accessible to a general audience. This is possible to implement using __shiny__ [@chang_shiny_2020]. The primary purposes of the interface would be to allow the user to interactively change the manip variable easily, and the interpolation step for more or less detailed views.


## Acknowledgments

This article was created in R, using __knitr__ [@xie_knitr_2020] and __rmarkdown__ [@allaire_rmarkdown_2020], with code generating the examples inline. The source files for this article be found at [github.com/nspyrison/spinifex_paper/](https://github.com/nspyrison/spinifex_paper/). The animated gifs can also be viewed at this site, and also in the supplementary material for this paper. The source code for the __spinifex__ package can be found at [github.com/nspyrison/spinifex/](https://github.com/nspyrison/spinifex/).

<!-- Adds a bib section at the end of every chapter -->

