<!-- Print bib before appendix -->
# Bibliography {-}

<div id="refs"></div>


<!-- ELSE use this at end to print html bib -->
<!-- `r if (knitr::is_html_output()) '# Bibliography {-}'` -->


# (APPENDIX) Appendix {-#ch-appendix}
<!-- # Appendix {#ch-appendix} -->

<!-- Does something for LaTeX output -->
\appendix


<!-- !!!LATEX TABLES WILL NOT WORK IN HTML, need to use kbl... -->
```{r}
require(tibble)
require(kableExtra)
require(magrittr)

glossary_df <- data.frame(
  Term = c("data",
           "observation",
           "variable",
           "Basis",
           "linear projection",
           "explanatory variables",
           "response variable"),
  Alias = c(" - ",
            "instance, item, case, row (of data)",
            "feature, column (of data)",
            "linear combination of variables, their orientations",
            "linear embedding",
            "independent-/input- variables, predictors, covariates",
            "predicted-/dependent-/target-/output- variable"),
  Terminology = NA ## Specified later for formatting
)

thesis_df <- tibble::tribble(
  ~Content,                                           ~Link,
  "thesis repository",                                "https://tinyurl.com/bddbs6sr", 
  #https://github.com/nspyrison/thesis_ns
  "thesis, pdf format",                               "https://tinyurl.com/5c54s7bf",
  #https://github.com/nspyrison/thesis_ns/blob/master/docs/thesis_ns.pdf
  "thesis, html format",                              "https://tinyurl.com/2p8m92bs",
  #https://nspyrison.github.io/thesis_ns/
  "Penguins radial tour, Figure \\@ref(fig:ch1fig3)", "https://vimeo.com/676723431",
  "Penguins grand tour, Figure \\@ref(fig:ch2fig4)",  "https://vimeo.com/676723441",
)

ch3_df <- tibble::tribble(
  ~Content,                                                   ~Link,
  "spinifex documentation website",                           "https://tinyurl.com/2p82e782",
  #"https://nspyrison.github.io/spinifex/",
  "Vignette: Getting started with spinifex",                  "https://tinyurl.com/2p98cfwh", 
  #"https://nspyrison.github.io/spinifex/articles/getting_started_with_spinifex.html",
  "Vignette: Ggproto api",                                    "https://tinyurl.com/2p8m6cmc",
  #"https://nspyrison.github.io/spinifex/articles/ggproto_api.html",
  "Fleas, radial tour animation, Figure \\@ref(fig:ch3fig3)", "https://tinyurl.com/3t6w7psf",
  #"https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/flea_radialtour_mvar5.gif"
  "Jet cluster, varying PC1",                                 "https://tinyurl.com/ye252yxb",
  #[PC1](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/jetcluster_manualtour_pc1.gif),
  "Jet cluster, varying PC2",                                 "https://tinyurl.com/ye252yxb",
  #[PC2](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/jetcluster_manualtour_pc2.gif),
  "Jet cluster, varying PC3, Figure \\@ref(fig:ch3fig5) -- least sensitive", "https://tinyurl.com/4wv3j92k",
  #least sensitive[PC3](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/jetcluster_manualtour_pc3.gif)
  "Jet cluster, varying PC4, Figure \\@ref(fig:ch3fig4) -- most sensitive",  "https://tinyurl.com/yckwdb5s",
  #most senitive[PC4](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/jetcluster_manualtour_pc4.gif)
  "DIS cluster, varying PC1",                                 "https://tinyurl.com/2wz377vh",
  #[PC1](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/discluster_manualtour_pc1.gif), 
  "DIS cluster, varying PC2, Figure \\@ref(fig:ch3fig7) -- least sensitive", "https://tinyurl.com/2hrxfdje",
  #least sensitive [PC2](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/discluster_manualtour_pc2.gif),
  "DIS cluster, varying PC3",                                 "https://tinyurl.com/ms683n2e",
  #[PC3](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/discluster_manualtour_pc3.gif),
  "DIS cluster, varying PC4",                                 "https://tinyurl.com/yhm244f7",
  #[PC4](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/discluster_manualtour_pc4.gif),
  "DIS cluster, varying PC5",                                 "https://tinyurl.com/4p5n3s77",
  #[PC5](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/discluster_manualtour_pc5.gif), 
  "DIS cluster, varying PC6, Figure \\@ref(fig:ch3fig6) -- most sensitive",  "https://tinyurl.com/mvcn2a8v",
  #most sensitive [PC6](https://github.com/nspyrison/spinifex_paper/blob/master/paper/gifs/discluster_manualtour_pc6.gif).
)

ch5_df <- tibble::tribble(
  ~Content,                                                           ~Link,
  "cheem documentation website",                                      "https://tinyurl.com/4avspber",
  #https://nspyrison.github.io/cheem/
  "Vignette: Getting started with cheem",                             "https://tinyurl.com/2p85sm4c",
  #https://nspyrison.github.io/cheem/articles/getting-started-with-cheem.html
  "cheem view applicaition",                                          "https://tinyurl.com/2hubrts7",
  #https://ebsmonash.shinyapps.io/cheem_initial/
  "Penguins animation, Figure \\@ref(fig:casepenguins)",              "https://vimeo.com/666431172",
  "Chocolates, animation, Figure \\@ref(fig:casechocolates)",         "https://vimeo.com/666431143",
  "Chocolates2, animation, Figure \\@ref(fig:casechocolatesinverse)", "https://vimeo.com/666431148",
  "FIFA 2020, animation, Figure \\@ref(fig:casefifa)",                "https://vimeo.com/666431163",
  "Ames house prices, animation, Figure \\@ref(fig:caseames)",        "https://vimeo.com/666431134",
)


## Set format and fonts for pdf vs html:
glossary_df$Terminology <- c(
    "$X_{n \\times p}$",
    "$X_{i \\times .} | i \\in [1, n]$",
    "$X_{. \\times j} | j \\in [1, p]$",
    "$A_{p \\times d} | A$  is orthonormal",
    "$Y_{n \\times d} = X_{n \\times p} \\times A_{p \\times d}$",
    "not used",
    "not used")
if(knitr::is_html_output()){
  fmt  <- "html"
  fnt  <- 12L
  fnt2 <- 12L
  esc  <- FALSE
}else{ ## Latex PDF
  fmt  <- "latex"
  fnt  <- 10L
  fnt2 <- 8L
  esc  <- FALSE ## latex needs \_ escaping in urls to use false.
  #glossary_df$Terminology <- c(rep(1,7))
  
  ## Use \@ for hardcoded rmarkdown , \\@ for format="html" or "simple" (html or pdf), and \\\\@ with format = "latex"
  # thesis_df$Content <- gsub("\\@", "\\\\@", thesis_df$Content)
  # ch3_df$Content    <- gsub("\\@", "\\\\@", ch3_df$Content)
  # ch5_df$Content    <- gsub("\\@", "\\\\@", ch5_df$Content)
}
```

# Glossary & terminology {#sec-glossary}

This thesis was written with statistics terms. This glossary helps bridge the language used here to be more accessible to machine learning audiences.


```{r glossary, fig.cap = "Use the caption arg in kable(), not this."}
# knitr::kable(glossary_df, format = "simple")

kableExtra::kbl(
  glossary_df, format = fmt,
  booktabs = TRUE, longtable = TRUE, linesep = "", escape = FALSE) %>%
kableExtra::kable_classic(font_size = fnt2)
```


# Supplementary material

Below collects links of documentation, code vignettes, and animations related to the content. The content for Chapter \@ref(ch-userstudy) continues with a bit of an extended analysis looking at the visual displays in the user study, the participant demographics, and a parallel analysis regressing of log time.


## Thesis

```{r fig.cap = "Use the caption arg in kable(), not this."}
knitr::kable(thesis_df, format = "simple")

# kableExtra::kbl(
#   thesis_df, format = fmt,
#   booktabs = TRUE, longtable = TRUE, linesep = "", escape = esc) %>%
#   kableExtra::kable_classic(font_size = fnt2)
```


## Chapter 3, spinifex links

```{r fig.cap = "Use the caption arg in kable(), not this."}
knitr::kable(ch3_df, format = "simple")

# kableExtra::kbl(
#   ch3_df, format = fmt,
#   booktabs = TRUE, longtable = TRUE, linesep = "", escape = esc) %>%
#   kableExtra::kable_classic(font_size = fnt2)
```


## Chapter 4, user study extended analysis {#sec:extanalysis}

<!-- segue to extended analysis -->
This section covers extended analysis. First, it illustrations of the different visuals are provided. Then, the participant demographics are covered. Lastly, a parallel modeling analysis on log response time is conducted.


### Illustrations of visual methods

Below illustrates the three visual methods evaluated in the user study. Data was collected from a __shiny__ application, and pre-rendered `gif` files were displayed based on the selected inputs.

```{r out.width="100%", fig.cap = "Examples of the application displays for PCA, grand tour, and radial tour."}
nsSafeIncGraphic("./figures/ch4_zapp_study_app_combined.PNG")
```

### Survey participant demographics {#sec:demographics}

The target population is relatively well-educated people, as linear projections may prove difficult for generalized consumption. Hence Prolific.co participants are restricted to those with an undergraduate degree (58,700 of the 150,400 users at the study time). From this cohort, 108 performed a complete study. Of these participants, 84 submitted the post-study survey, represented in the following heatmap. All participants were compensated for their time at \pounds 7.50 per hour, with a mean time of about 16 minutes. Figure \@ref(fig:figSurveyDemographics) shows a heat map of the demographics for these 84 participants.

```{r, figSurveyDemographics, out.width="100%", fig.cap = "Heatmaps of survey participant demographics; counts of age group by completed education as faceted across preferred pronoun. Our sample tended to be between 18 and 35 years of age with an undergraduate or graduate degree."}
nsSafeIncGraphic("./figures/ch4_zapp_fig1_survey_demograpics.pdf")
```


### Response time {#sec:responsetime}

<!-- Time as secondary interest, Y2 -->
As a secondary explanatory variable, response time is considered. Response time is first log-transformed to remove its right skew. The same modeling procedure is repeated for this response. 1) Compare the performance of a battery of all additive and multiplicative models. Table \@ref(tab:timeCompTbl) shows the higher level performance of these models over increasing model complexity. 2) Select the model with the same effect terms, $\alpha \times \beta + \gamma + \delta$, with relatively high conditional $R^2$ without becoming overly complex from interaction. The coefficients of this model are displayed in Table \@ref(tab:timeCoefTbl).

<!-- Y2 model comparisons, continue to use ABcd -->
```{r timeCompTbl, fig.cap = "Use the caption arg in kable(), not this."}
## Set format and fonts for pdf vs html:
if(knitr::is_html_output()){
  fmt      <- "html"
  fnt      <- 12L
  mod_comp <- readRDS("./figures/ch4_tab1_model_comp_ls_html.rds")
  ## Escape HTML formatting of * multiplication
  mod_comp$modelComp_MarksByEval$`Fixed effects` <-
      gsub("[*]", "\\\\*", mod_comp$modelComp_MarksByEval$`Fixed effects`)
  mod_comp$modelComp_TimeByEval$`Fixed effects`  <-
      gsub("[*]", "\\\\*", mod_comp$modelComp_TimeByEval$`Fixed effects`)
}else{
  fmt      <- "latex"
  fnt      <- 10L
  mod_comp <- readRDS("./figures/ch4_tab1_model_comp_ls_latex.rds")
}
esc_comp <- FALSE ## Comparisons, with bolding 
esc_coef <- TRUE  ## Coefficients, with % and *

## Y2, log time comparison table
kableExtra::kbl(
  mod_comp[[2]], format = fmt, align = c("l", rep("l", 2), rep("c", 5)),
  booktabs = TRUE, linesep = "", escape = esc_comp,
  caption = "Model performance regressing on log response time [seconds], $\\widehat{Y_2}$ random effect models. Conditional $R^2$ includes the random effects, while marginal does not. The model $\\alpha \\times \\beta + \\gamma + \\delta$ model is selected to examine further as it has relatively high marginal $R^2$ while having much less complexity than the complete interaction model.") %>%
  kableExtra::kable_classic(font_size = fnt)
```

```{r timeCoefTbl, fig.cap = "Use the caption arg in kable(), not this."}
## Y2, log time coeffiecients
kableExtra::kbl(
  mod_coef[[2]], booktabs = TRUE, linesep = "", format = fmt, escape = esc_coef,
  caption = "Model coefficients for log response time [seconds] $\\widehat{Y_2} = \\alpha \\times \\beta + \\gamma + \\delta$, with factor = pca, location = 0/100\\%, shape = EEE, and dim = 4 held as baselines. Location = 50/50\\% is the fixed term with the strongest evidence and takes less time. In contrast, the interaction term location = 50/50\\%:shape = EEV has the most evidence and takes much longer on average.") %>%
  kableExtra::pack_rows("Visual", 2, 3) %>%
  kableExtra::pack_rows("Fixed effects", 4, 8) %>%
  kableExtra::pack_rows("Interactions", 9, 12) %>%
  kableExtra::kable_classic(font_size = fnt)
```


## Chapter 5, cheem links

```{r fig.cap = "Use the caption arg in kable(), not this."}
knitr::kable(ch5_df, format = "simple")
```


<!-- ===== Removed, random effect range analysis ======= -->
<!-- #### Random effect ranges {-} -->

<!-- <!-- Random effect terms specify source of the error -->
<!-- The random effect terms further clarify the source of the error. Below is a comparison the effect ranges attributed to the participant and to the simulations next to their marginal effect on the response, a sort of upper bound of the error they could explain. This was performed for the models regressing accuracy and then log response time. -->

<!-- <!-- Random effects vs Mean Mark CI by participant and sim -->
<!-- The residual plots have no noticeable nonlinear trends and contain striped patterns as an artifact from regressing on discrete variables. Figure \@ref(fig:figEffectRange) illustrates (T) the effect size of the random terms participant and simulation, or more accurately, the 95\% CI from Gelman simulation of their posterior distribution. The effect size of the participant is much larger than simulation. The most extreme participants are statistically significant at $\alpha = .95$, while none of the simulation effects significantly deviate from the null of having no effect size on the marks. In comparison, (B) 95\% confidence intervals participation and simulation mean accuracy, respectively. -->

<!-- Residual plots have no noticeable nonlinear trends and contain striped patterns as an artifact from regressing on discrete variables. Figure \@ref(fig:figEffectRange) illustrates (T) the effect size of the random terms participant and simulation, or more accurately, the 95\% CI from Gelman simulation of their posterior distribution. The effect size of the participant is much larger than simulation. The most extreme participants are statistically significant at $\alpha = .95$, while none of the simulation effects significantly deviate from the null of having no effect size on the marks. In comparison, (B) 95\% confidence intervals participation and simulation mean accuracy, respectively. -->

<!-- ```{r figEffectRange, out.width="100%", fig.show="asis", fig.cap="Accuracy model: (T) Estimated effect ranges of the random effect terms participant and data simulation of the accuracy model, $\\widehat{Y_1} = \\alpha \\times \\beta + \\gamma + \\delta$. Confidence intervals are created with Gelman simulation on the effect posterior distributions. The effect size of the participant is relatively large, with several significant extrema. None of the simulations deviate significantly. (B) The ordered distributions of the CI of mean marks follow the same general pattern and give the additional context of how much variation is in the data, an upper limit to the effect range. The effect ranges capture about two-thirds of the range of the data without the model. All intervals for $\\alpha = .95$ confidence."} -->
<!-- nsSafeIncGraphic("./figures/ch4_zapp_fig2_effect_range.pdf") -->
<!-- ``` -->

<!-- Similarly, figure \@ref(fig:figTeffectRange) shows the Gelman simulations and marginal effects of the simulation and participants for the model with the same terms regressing on log response time. -->

<!-- ```{r figTeffectRange, out.width="100%", fig.show="asis", fig.cap = "Log response time model: (T) The effect ranges of Gelman resimulation on posterior distributions for the time model, $\\widehat{Y_2} = \\alpha \\times \\beta + \\gamma + \\delta$. These show the magnitude and distributions of particular participants and simulations. Simulation has a relatively small effect on response time. (B) Confidence intervals for mean log time by participant and simulation. The marginal density shows that the response times are left-skewed after log transformation. Interpreting back to linear time there is quite the spread of response times: $e^{1} = 2.7$, $e^{2.75} = 15.6$, $e^{3.75} = 42.5$ seconds. Of the simulations on the right, the bottom has a large variation in response time, relative to the effect ranges which means that the variation is explained in the terms of the model and not by the simulation itself."} -->
<!-- nsSafeIncGraphic("./figures/ch4_zapp_fig3_T_effect_range.pdf") -->
<!-- ``` -->


<!-- ## Accompanying radial tour application {#sec:spinifex} -->

<!-- An accompanying application illustrates the radial tour. The __R__ package, __spinifex__, [@spyrison_spinifex_2020] is an open-source and now contains a __shiny__ [@chang_shiny_2021] application allowing users to apply various preprocessing tasks and interactively explore their data via interactive radial tour. Example datasets are provided with the ability to upload data. The `html` widget produced is a more interactive variant relative to the one used in the user study. Screen captures and more details are provided in the appendix. Run the following __R__ code will run the application locally. -->

<!-- ```{r getting_started, eval=FALSE, echo=TRUE} -->
<!-- install.packages("spinifex", dependencies = TRUE) -->
<!-- spinifex::run_app() -->
<!-- ``` -->

<!-- ```{r radialTourAppPg1, out.width="100%", fig.show="asis", fig.cap = "Process data tab, interactively loads or select data, check which variables to project, and optionally scale columns by standard deviation."} -->
<!-- nsSafeIncGraphic("./figures/ch4_zapp_fig4_app_pg1.PNG") -->
<!-- ``` -->

<!-- In the initial tab, Figure \@ref(fig:radialTourAppPg1), users upload their own (.csv, .rds, or .rda) data or select from predefined data sets. The numeric columns appear as a list of variables to include in the projection. Below that, a line displays whether or not missing rows were removed. Scaling by standard deviation is included by default, as this is a common transformation used to explore linear projections of spaces. Summaries of the raw data and processed numeric data are displayed to illustrate how the data was read and its transformation. -->

<!-- ```{r radialTourAppPg2, out.width="100%", fig.show="asis", fig.cap = "Radial tour tab, interactively create radial tours, changing the manipulation variable, color, or shape of the resulting manual tour. Here, the palmer penguins data is being explored, bill length was selected to manipulate as it is the only variable separating the green cluster from the orange. By mapping shape to island of observation, the green species can be noted to live on all three islands, while the other species live on only one island."} -->
<!-- nsSafeIncGraphic("./figures/ch4_zapp_fig5_app_pg2.PNG") -->
<!-- ``` -->

<!-- The second tab, Figure \@ref(fig:radialTourAppPg2) contains interaction for selecting the manipulation variable, and non-numeric columns can be used to change the color and shape of the data points in the projection. The radial tour is created in real-time, animated as an interactive __plotly__ `html` widget. The application offers users a fast, intuitive introduction elucidating what the radial tour does and some of the features offered. -->






<!-- ===== OLD notes about tours, index, interpolation, and applications =====
### Tour taxonomy; path generation {#sec:path_generation}

A fundamental aspect of tours is the path of rotation. There are four primary distinctions of tour path generation [@buja_computational_2005]: random choice, data-driven, precomputed choice, and manual control.

* Random choice, *grand tour*, constrained random walks $p$-space. Paths are constrained for changes in direction small enough to maintain continuity and aid in user comprehension
    + torus-surface [@asimov_grand_1985]
    + at-random [@asimov_grand_1985]
    + random-walk [@asimov_grand_1985]
    + *local tour* [@wickham_tourr_2011], a sort of grand tour on a leash, such that it goes to a nearby random projection before returning to the original position and iterating to a new nearby projection.
* _guided tour_, optimizes an objective function/index of the projection using projection pursuit (PP) [@hurley_analyzing_1990], including the following implemented indexes:
    + holes [@cook_projection_1993] - moves points away from the center.
    + cmass [@cook_projection_1993] - moves points toward the center.
    + lda [@lee_projection_2005] - linear discriminant analysis, seeks a projection where 2 or more classes are most separated.
    + pda [@lee_projection_2010] - penalized discriminant analysis for use in highly correlated variables when classification is needed.
    + convex [@laa_using_2019] - the ratio of the area of convex and alpha hulls.
    + skinny [@laa_using_2019] - the ratio of the perimeter distance to the area of the alpha hull.
    + stringy [@laa_using_2019] - based on the minimum spanning tree (MST), the diameter of the MST over the length of the MST.
    + dcor2D [@grimm_mbgraphic:_2017; @laa_using_2019] - distance correlation that finds linear and nonlinear dependencies between variables.
    + splines2D [@grimm_mbgraphic:_2017; @laa_using_2019] - measure of nonlinear dependence by fitting spline models.
    + other user-defined objective indices can be applied to the framework provided in the *tourr* package @wickham_tourr_2011.
    + Another data-drive tour is the *dependence tour*, a combination of $n$ independent 1D tours. A vector describes the axis each variable will be displayed on. for example $c(1, 1, 2, 2)$ is a 4- to 2D tour with the first 2 variables on the first axis, and the remaining on the second.
        - *correlation tour* [@buja_data_1987], a special case of the dependence tour, analogous to canonical correlation analysis.
* Precomputed choice, *planned tour*, in which the path has already been generated or defined.
    + *little tour* [@mcdonald_interactive_1982], where every permutation of variables is stepped through in order, analogous to brute-force or exhaustive search.
    + a saved path of any other tour, typically an array of basis targets to interpolate between.
* Manual control, *manual tour*, a constrained rotation on selected manipulation variable and magnitude [@cook_manual_1997]. Typically used to explore the local area after identifying an interesting feature, perhaps via guided tour.
    + radial, fix angle of contribution, change magnitude radially.
    + oblique, along an oblique path of a dragged cursor


### Interpolation

NOTE: Search 'Interpolator' in Wickham’s tourr paper
After target bases are identified, the frames in-between need to be filled in. There are several methods to do so:

* Geodesic - via Gram-Schmidt process
* Givens rotations
* Householder reflections
.

### Tour software

Tours have yet to be widely adopted, due in part, to the fact that print and static pdf output does not accommodate dynamic viewing. Conceptual abstraction and technically density have also hampered user growth. Due to low levels of adoption and the rapid advancement of technology support and maintenance of such implementations give them a particularly short life span. Despite the small user base, there have been a fair number of tour implementations, including:
<!-- See Wickham’s thesis and C2 paper for partial lists. 

* spinifex [github.com/nspyrison/spinifex](https://github.com/nspyrison/spinifex) -- R package, all platforms.
* tourr [@wickham_tourr_2011] -- R package, all platforms.
* CyrstalVision [@wegman_visual_2003] -- for Windows.
* GGobi [@swayne_ggobi:_2003] -- for Linux and Windows.
* DAVIS [@huh_davis:_2002] -- Java based, with GUI.
* ORCA [@sutherland_orca:_2000] -- Extensible toolkit built in Java.
* VRGobi [@nelson_xgobi_1999] -- for use with the C2, tours in stereoscopic 3D displays.
* ExplorN [@carr_explorn:_1996] -- for SGI Unix.
* ExploRe [@hardle_xplore:_1995]
* XGobi [@swayne_xgobi:_1991] -- for Linux, Unix, and Windows (via emulation).
* XLispStat [@tierney_lisp-stat:_1990] -- for Unix and Windows.
* Explor4 [@carr_explor4:_1988] -- Four-dimensional data using stereo-ray glyphs.
* Prim-9 [@asimov_grand_1985;@fisherkeller_prim-9:_1974] -- on an internal operating system.
* liminal
* cheem
-->

