# Appendix {#ch-appendix}

<!-- Does something for LaTeX output -->
\appendix


## Accompanying tool: radial tour application {#sec:spinifex}

An accompanying application illustrates the radial tour. The __R__ package, __spinifex__, [@spyrison_spinifex_2020] is an open-source and now contains a __shiny__ [@chang_shiny_2021] application allowing users to apply various preprocessing tasks and interactively explore their data via interactive radial tour. Example datasets are provided with the ability to upload data. The `html` widget produced is a more interactive variant relative to the one used in the user study. Screen captures and more details are provided in the appendix. Run the following __R__ code will run the application locally.

```{r getting_started, eval=FALSE, echo=TRUE}
## Download:
install.packages("spinifex", dependencies = TRUE)
## Run interactive app demonstrating the radial tour:
spinifex::run_app()
```

```{r radialTourAppPg1, out.width="100%", fig.show="asis", fig.cap = "Process data tab, interactively loads or select data, check which variables to project, and optionally scale columns by standard deviation."}
nsSafeIncGraphic("./figures/ch4_zapp_fig4_app_pg1.PNG")
```

In the initial tab, Figure \@ref(fig:radialTourAppPg1), users upload their own (.csv, .rds, or .rda) data or select from predefined data sets. The numeric columns appear as a list of variables to include in the projection. Below that, a line displays whether or not missing rows were removed. Scaling by standard deviation is included by default, as this is a common transformation used to explore linear projections of spaces. Summaries of the raw data and processed numeric data are displayed to illustrate how the data was read and its transformation.

```{r radialTourAppPg2, out.width="100%", fig.show="asis", fig.cap = "Radial tour tab, interactively create radial tours, changing the manipulation variable, color, or shape of the resulting manual tour. Here, the palmer penguins data is being explored, bill length was selected to manipulate as it is the only variable separating the green cluster from the orange. By mapping shape to island of observation, the green species can be noted to live on all three islands, while the other species live on only one island."}
nsSafeIncGraphic("./figures/ch4_zapp_fig5_app_pg2.PNG")
```

The second tab, Figure \@ref(fig:radialTourAppPg2) contains interaction for selecting the manipulation variable, and non-numeric columns can be used to change the color and shape of the data points in the projection. The radial tour is created in real-time, animated as an interactive __plotly__ `html` widget. The application offers users a fast, intuitive introduction elucidating what the radial tour does and some of the features offered.


## User study extended analysis {#sec:appendix}

<!-- segue to extended analysis -->
This section covers peripheral and extended analysis. First, the demographics of the participants is covered. Then a parallel modeling analysis on log response time is conducted. Lastly, the effect ranges and marginal effects of the random effect of the participants and data simulation are examined.


### Survey participant demographics {-}

The target population is relatively well-educated people, as linear projections may prove difficult for generalized consumption. Hence Prolific.co participants are restricted to those with an undergraduate degree (58,700 of the 150,400 users at the study time). From this cohort, 108 performed a complete study. Of these participants, 84 submitted the post-study survey, represented in the following heatmap. All participants were compensated for their time at \pounds 7.50 per hour, with a mean time of about 16 minutes. Figure \@ref(fig:figSurveyDemographics) shows a heat map of the demographics for these 84 participants.

```{r, figSurveyDemographics, out.width="100%", fig.cap = "Heatmaps of survey participant demographics; counts of age group by completed education as faceted across preferred pronoun. Our sample tended to be between 18 and 35 years of age with an undergraduate or graduate degree."}
nsSafeIncGraphic("./figures/ch4_zapp_fig1_survey_demograpics.pdf")
```


### Response time regression

<!-- Time as secondary interest, Y2 -->
As a secondary explanatory variable response time is considered. Response time is first log transformed to remove its right skew. The same modeling procedure is repeated for this response. 1) Compare the performance of a battery of all additive and multiplicative models. Table \@ref(tab:timeCompTbl) shows the higher level performance of these models over increasing model complexity. 2) Select the model with the same effect terms, $\alpha \cdot \beta + \gamma + \delta$, and examine its coefficients, displayed in Table \@ref(tab:timeCoefTbl).

<!-- Y2 model comparisons, continue to use ABcd -->
```{r timeCompTbl, fig.cap = "Use the caption arg in kable(), not this."}
kableExtra::kbl(
  mod_comp[[2]], format = fmt, align = c("l", rep("l", 2), rep("c", 5)),
  booktabs = TRUE, linesep = "", row.names = FALSE, escape = esc_comp,
  caption = "Model performance regressing on log response time [seconds], $\\widehat{Y_2}$ random effect models, where each includes random effect terms for participants and simulations. There is a simplar trade-off where AIC/BIC prefer the simplest factor model, while $R^2$ and RMSE are the largest in the full multiplicative model. The model $\\alpha \\cdot \\beta + \\gamma + \\delta$ model is selected to examine further as it has relatively high marginal $R^2$ while having much less complexity than the complete interaction model. Conditional $R^2$ includes the random effects, while marginal does not.") %>%
  kableExtra::kable_classic(font_size = fnt) #bootstrap_options = "striped", 
```

<!-- Y2 coeffiecients -->
```{r timeCoefTbl, fig.cap = "Use the caption arg in kable(), not this."}
kableExtra::kbl(
  mod_coef[[2]], booktabs = TRUE, linesep = "", format = fmt, escape = esc_coef,
  caption = "Model coefficients for log response time [seconds] $\\widehat{Y_2} = \\alpha \\cdot \\beta + \\gamma + \\delta$, with factor = pca, location = 0/100\\%, shape = EEE, and dim = 4 held as baselines. Location = 50/50\\% is the fixed term with the strongest evidence and takes less time. In contrast, the interaction term location = 50/50\\%:shape = EEV has the most evidence and takes much longer on average.") %>%
  kableExtra::pack_rows("Factor", 2, 3) %>%
  kableExtra::pack_rows("Fixed effects", 4, 8) %>%
  kableExtra::pack_rows("Interactions", 9, 12) %>%
  kableExtra::kable_classic(font_size = fnt) #bootstrap_options = "striped", 
```


### Random effect ranges {-}

<!-- Random effect terms specify source of the error -->
The random effect terms further clarify the source of the error. Below is a comparison the effect ranges attributed to the participant and to the simulations next to their marginal effect on the response, a sort of upper bound of the error they could explain. This was performed for the models regressing accuracy and then log response time.

<!-- Random effects vs Mean Mark CI by participant and sim -->
The residual plots have no noticeable nonlinear trends and contain striped patterns as an artifact from regressing on discrete variables. Figure \@ref(fig:figEffectRange) illustrates (T) the effect size of the random terms participant and simulation, or more accurately, the 95\% CI from Gelman simulation of their posterior distribution. The effect size of the participant is much larger than simulation. The most extreme participants are statistically significant at $\alpha = .95$, while none of the simulation effects significantly deviate from the null of having no effect size on the marks. In comparison, (B) 95\% confidence intervals participation and simulation mean accuracy, respectively.

Residual plots have no noticeable nonlinear trends and contain striped patterns as an artifact from regressing on discrete variables. Figure \@ref(fig:figEffectRange) illustrates (T) the effect size of the random terms participant and simulation, or more accurately, the 95\% CI from Gelman simulation of their posterior distribution. The effect size of the participant is much larger than simulation. The most extreme participants are statistically significant at $\alpha = .95$, while none of the simulation effects significantly deviate from the null of having no effect size on the marks. In comparison, (B) 95\% confidence intervals participation and simulation mean accuracy, respectively.

```{r figEffectRange, out.width="100%", fig.show="asis", fig.cap="Accuracy model: (T) Estimated effect ranges of the random effect terms participant and data simulation of the accuracy model, $\\widehat{Y_1} = \\alpha \\cdot \\beta + \\gamma + \\delta$. Confidence intervals are created with Gelman simulation on the effect posterior distributions. The effect size of the participant is relatively large, with several significant extrema. None of the simulations deviate significantly. (B) The ordered distributions of the CI of mean marks follow the same general pattern and give the additional context of how much variation is in the data, an upper limit to the effect range. The effect ranges capture about two-thirds of the range of the data without the model. All intervals for $\\alpha = .95$ confidence."}
nsSafeIncGraphic("./figures/ch4_zapp_fig2_effect_range.pdf")
```

Similarly, figure \@ref(fig:figTeffectRange) shows the Gelman simulations and marginal effects of the simulation and participants for the model with the same terms regressing on log response time.

```{r figTeffectRange, out.width="100%", fig.show="asis", fig.cap = "Log response time model: (T) The effect ranges of Gelman resimulation on posterior distributions for the time model, $\\widehat{Y_2} = \\alpha \\cdot \\beta + \\gamma + \\delta$. These show the magnitude and distributions of particular participants and simulations. Simulation has a relatively small effect on response time. (B) Confidence intervals for mean log time by participant and simulation. The marginal density shows that the response times are left-skewed after log transformation. Interpreting back to linear time there is quite the spread of response times: $e^{1} = 2.7$, $e^{2.75} = 15.6$, $e^{3.75} = 42.5$ seconds. Of the simulations on the right, the bottom has a large variation in response time, relative to the effect ranges which means that the variation is explained in the terms of the model and not by the simulation itself."}
nsSafeIncGraphic("./figures/ch4_zapp_fig3_T_effect_range.pdf")
```


<!-- !!!LATEX TABLES WILL NOT WORK IN HTML, need conditional kbl formating... -->
```{r}
glossary_df <- data.frame(
  Term = c("observation", "variable", "data", "basis", "projection", "tour",
           "Intrinsic data dimensionality", "Nonlinear model"),
  Notation = c(
    "$n$", "$p$", "$X_{n \\times p}$", "$A_{p \\times d}$",
    "$Y_{n \\times d} = X_{n \\times p} \\cdot A_{p \\times d}$", "", "$iid$", ""),
  Alias = c(
    "instance, item, row (of data)", "feature, column (of data)",
    "", "", "embedding", "", "", "black-box model"),
  Description = c(
    "A unit being measured across the variables",
    "A measure taken across all observations",
    "We dicuss complete quantitative matrices",
    "Of linear projections, linear component mapping of $p$ down to $d$ dimensions",
    "The resulting space the data multiplied by the basis",
    "A dynamic linear projection animated over small changes in the projection basis",
    "The minimal dimensionality data can be represented with",
    "A model with numerous or interactive terms that cause an interpretabilty issue")
)

vimeo_link_df <- data.frame(
  Reference = c(
    "Figure \\@ref(fig:ch1fig3)", "Figure \\@ref(fig:ch2fig4)", 
    "Figure \\@ref(fig:casepenguins)", "Figure \\@ref(fig:casechocolates)", 
    "Figure \\@ref(fig:casechocolatesinverse)", "Figure \\@ref(fig:casefifa)", 
    "Figure \\@ref(fig:caseames)"),
  Description = c(
    "radial tour, penguins", "grand tour, penguins", "cheem,  penguins", 
    "cheem,  chocolates", "cheem,  chocolates (inverse)", "cheem, fifa", 
    "cheem, ames housing 2018"),
  Link = c("[vimeo.com/676723431](https://vimeo.com/676723431)",
           "[vimeo.com/676723441](https://vimeo.com/676723441)",
           "[vimeo.com/666431172](https://vimeo.com/666431172)",
           "[vimeo.com/666431143](https://vimeo.com/666431143)",
           "[vimeo.com/666431148](https://vimeo.com/666431148)",
           "[vimeo.com/666431163](https://vimeo.com/666431163)",
           "[vimeo.com/666431134](https://vimeo.com/666431134)")
)

supp_mats_df <- data.frame(
  Context = c(
    #"spinifex", 
    "spinifex", "spinifex", 
    #"cheem", 
    "cheem", "thesis", "thesis", "thesis"),
  Description = c(
    #"CRAN Link", 
    "vignette, Getting started with spinifex", 
    "vignette, Ggproto api", 
    #"CRAN Link", 
    "vignette, Getting started with cheem", 
    "repository", "thesis, pdf format", "thesis, html format"),
  Link = c(
    #"[CRAN.R-project.org/package=spinifex](https://CRAN.R-project.org/package=spinifex)",
    "[nspyrison.github.io/spinifex/articles/getting-started-with-spinifex.html](https://nspyrison.github.io/spinifex/articles/getting-started-with-spinifex.html)",
    "[nspyrison.github.io/spinifex/articles/ggproto-api.html](https://nspyrison.github.io/spinifex/articles/ggproto-api.html)",
    #"[CRAN.R-project.org/package=cheem](https://CRAN.R-project.org/package=cheem)",
    "[nspyrison.github.io/cheem/articles/getting-started-with-cheem.html](https://nspyrison.github.io/cheem/articles/getting-started-with-cheem.html)",
    "[github.com/nspyrison/thesis_ns](https://github.com/nspyrison/thesis_ns)",
    "[github.com/nspyrison/thesis_ns/blob/master/docs/thesis_ns.pdf](https://github.com/nspyrison/thesis_ns/blob/master/docs/thesis_ns.pdf)",
    "[https://nspyrison.github.io/thesis_ns/](https://nspyrison.github.io/thesis_ns/)")
)

## Set format and fonts for pdf vs html:
if(knitr::is_html_output()){
  fmt <- "html"
  fnt <- 12L
  esc <- FALSE ## HTML likes FALSE in vimeolinks
}else{
  fmt <- "latex"
  fnt <- 10L
  esc <- TRUE #FALSE ## latex needs \ escaping on urls to use false.
  #supp_mats_df$Link <- gsub("_", "\_", supp_mats_df$Link) 
}
```

## Glossary {#sec-glossary}

```{r glossary, fig.cap = "Use the caption arg in kable(), not this."}
kableExtra::kbl(
    glossary_df, format = fmt, align = rep("l", 4),
    booktabs = TRUE, longtable = TRUE, linesep = "", escape = esc,
    caption = "Glossary of terms") %>%
  kableExtra::kable_classic(font_size = fnt)
```


## Animated tour links

```{r vimeolinks, fig.cap = "Use the caption arg in kable(), not this."}
kableExtra::kbl(
    vimeo_link_df, format = fmt, align = rep("l", 3),
    booktabs = TRUE, longtable = TRUE, linesep = "", escape = esc,
    caption = "Animated tour links in this thesis") %>%
  kableExtra::kable_classic(font_size = fnt)
```


## Supplementary material

```{r suppmats, fig.cap = "Use the caption arg in kable(), not this."}
kableExtra::kbl(
    supp_mats_df, format = fmt, align = rep("l", 3),
    booktabs = TRUE, longtable = TRUE, linesep = "", escape = esc,
    caption = "Supplementary material") %>%
  kableExtra::kable_classic(font_size = fnt)
```


<!---
## Notation {#sec:notation}

Tour notation varies across articles and authors. In my work, I use the following:

* $n$, number of observations in the data.
* $p$, number of numeric variables, the dimensionality of data space.
* $d$, the dimensionality of projection space.
* $\textbf{X}_{[n,~p]}$, a data matrix in variable-space, $\textbf{X} \in \mathbb{R}^{p}$. Typically centered, scaled, and optionally sphered.
* $\textbf{A}_{[p,~d]}$, an orthonormal matrix (columns are independant, at right angles with eachother and each normalized to a length of 1). This is the linear combination of variables from the orignal space to a lower embedding space. That is, the orientation of the variables, mapping from   $p-$ to $d-$space.
* $\textbf{Y}_{[n,~d]}$, projected data matrix in projection-space, $\textbf{Y} \in \mathbb{R}^{d}$.
    * In chapter 
* Reference axes, a display showing how each variables coefficient(s) contribute to a projection. Either on its own axis (1D) or relative to a unit circle (2D).
* Geometric objects are referred to in generalized dimensions; the use of the term plane is not necessarily a 2D surface, but a hyperplane in the arbitrary dimensions of the projection space.
-->

<!-- 
## Data visualization terminology {#sec:terminology}

* 2D - representation of data in 2 dimensions, without the use of depth perception cues and minimal aesthetic mapping (such as color, size, and height) to data points.
* 2.5D - following the definition given in @ware_designing_2000: visualizations that are essentially 2D but select depth cues are used to provide some suggestion of 3D. However, the term 2.5D is commonly used for several meanings *due to the ambiguous use of 2.5D, this document errs on the side stating 3D with descriptions of depth cues used*.
* 3D - visualizations of 3 dimensions with liberal use of depth cues unless otherwise qualified.
* Depth perception cues - an indication that indicates the depth to an observer, including:
    * linear perspective - the property of parallel lines converging on a vanishing point.
    * aerial perspective - objects that far away have lower contrast and color saturation due to light scattering in the atmosphere.
    * occultation (or interposition) - where closer objects partially block the view of further objects.
    * motion perspective/parallax - closer objects, move across the field of view faster than further objects.
    * accommodation - the change of focal length due to change in the shape of the eye. Effective for distances of less than 2 meters.
    * binocular stereopsis/disparity - the use of 2 images of slightly varied angles from the horizontal distance of the eyes. The disparity for distant objects is small, but it is significant for nearby objects.
    * binocular convergence - The ocular-motor cue due to stereopsis focusing on the same objects. Convergence is effective for distances up to 10 meters.
* Virtual reality (VR) - an immersive experience of computer-generated sensory input.
* Augmented reality (AR) - view of physical spaces with augmenting/ supplementing sensory input of information.
* Mixed reality (MR) - a mix of physical and virtual realities with objects from both interacting in real time. This differs from AR by the flow of interaction; AR augments physical reality while MR has reciprocating interactions.
* Extended reality (XR) - any degree of virtual, augmented, or mixed reality.
* Scatterplot matrices (SPLOMs) - matrix display of pair-wise 2D scatterplots, sometimes with 1D density on the diagonal.
* Human-in-the-loop - any model that requires human interaction [@karwowski_international_2006]. 
-->

<!--
### Tour taxonomy; path generation {#sec:path_generation}

A fundamental aspect of tours is the path of rotation. There are four primary distinctions of tour path generation [@buja_computational_2005]: random choice, data-driven, precomputed choice, and manual control.

* Random choice, *grand tour*, constrained random walks $p$-space. Paths are constrained for changes in direction small enough to maintain continuity and aid in user comprehension
    + torus-surface [@asimov_grand_1985]
    + at-random [@asimov_grand_1985]
    + random-walk [@asimov_grand_1985]
    + *local tour* [@wickham_tourr_2011], a sort of grand tour on a leash, such that it goes to a nearby random projection before returning to the original position and iterating to a new nearby projection.
* Data-driven, *guided tour*, optimizing some objective function/index within the projection-space, called projection pursuit (PP) [@hurley_analyzing_1990], including the following indexes:
    + holes [@cook_projection_1993] - moves points away from the center.
    + cmass [@cook_projection_1993] - moves points toward the center.
    + lda [@lee_projection_2005] - linear discriminant analysis, seeks a projection where 2 or more classes are most separated.
    + pda [@lee_projection_2010] - penalized discriminant analysis for use in highly correlated variables when classification is needed.
    + convex [@laa_using_2019] - the ratio of the area of convex and alpha hulls.
    + skinny [@laa_using_2019] - the ratio of the perimeter distance to the area of the alpha hull.
    + stringy [@laa_using_2019] - based on the minimum spanning tree (MST), the diameter of the MST over the length of the MST.
    + dcor2D [@grimm_mbgraphic:_2017; @laa_using_2019] - distance correlation that finds linear and nonlinear dependencies between variables.
    + splines2D [@grimm_mbgraphic:_2017; @laa_using_2019] - measure of nonlinear dependence by fitting spline models.
    + other user-defined objective indices can be applied to the framework provided in the *tourr* package @wickham_tourr_2011.
    + Another data-drive tour is the *dependence tour*, a combination of $n$ independent 1D tours. A vector describes the axis each variable will be displayed on. for example $c(1, 1, 2, 2)$ is a 4- to 2D tour with the first 2 variables on the first axis, and the remaining on the second.
        - *correlation tour* [@buja_data_1987], a special case of the dependence tour, analogous to canonical correlation analysis.
* Precomputed choice, *planned tour*, in which the path has already been generated or defined.
    + *little tour* [@mcdonald_interactive_1982], where every permutation of variables is stepped through in order, analogous to brute-force or exhaustive search.
    + a saved path of any other tour, typically an array of basis targets to interpolate between.
* Manual control, *manual tour*, a constrained rotation on selected manipulation variable and magnitude [@cook_manual_1997]. Typically used to explore the local area after identifying an interesting feature, perhaps via guided tour.
    + radial, fix angle of contribution, change magnetude radially.
    + oblique, along an oblique path of a draged cursor
    + 


### Interpolation

NOTE: Search 'Interpolator' in Wickham’s tourr paper
After target bases are identified, the frames in-between need to be filled in. There are several methods to do so:

* Geodesic - via Gram-Schmidt process
* Givens rotations
* Householder reflections
.

### Tour software

Tours have yet to be widely adopted, due in part, to the fact that print and static pdf output does not accommodate dynamic viewing. Conceptual abstraction and technically density have also hampered user growth. Due to low levels of adoption and the rapid advancement of technology support and maintenance of such implementations give them a particularly short life span. Despite the small user base, there have been a fair number of tour implementations, including:
<!-- See Wickham’s thesis and C2 paper for partial lists. 

* spinifex [github.com/nspyrison/spinifex](https://github.com/nspyrison/spinifex) -- R package, all platforms.
* tourr [@wickham_tourr_2011] -- R package, all platforms.
* CyrstalVision [@wegman_visual_2003] -- for Windows.
* GGobi [@swayne_ggobi:_2003] -- for Linux and Windows.
* DAVIS [@huh_davis:_2002] -- Java based, with GUI.
* ORCA [@sutherland_orca:_2000] -- Extensible toolkit built in Java.
* VRGobi [@nelson_xgobi_1999] -- for use with the C2, tours in stereoscopic 3D displays.
* ExplorN [@carr_explorn:_1996] -- for SGI Unix.
* ExploRe [@hardle_xplore:_1995]
* XGobi [@swayne_xgobi:_1991] -- for Linux, Unix, and Windows (via emulation).
* XLispStat [@tierney_lisp-stat:_1990] -- for Unix and Windows.
* Explor4 [@carr_explor4:_1988] -- Four-dimensional data using stereo-ray glyphs.
* Prim-9 [@asimov_grand_1985;@fisherkeller_prim-9:_1974] -- on an internal operating system.
* liminal
* cheem
-->

