---
chapter: 3
knit: "bookdown::render_book"
---

```{r setup_03, include=FALSE,}
knitr::opts_chunk$set(
  echo = TRUE, 
  fig.width = 6, 
  fig.asp = 1/1.61, 
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "",
  cache = FALSE
)
```

# *spinifex*: An R package that provides manual rotations in high-dimensions {#ch:spinifex}

## Abstract

Touring techniques offer a great opportunity for data-space visualizations of ($\textbf{X} \in \mathbb{R}^p,~p > 3$) multivariate data sets. This paper discusses the *R* package *spinifex*, which adds support for the manual tour, which is particularly usefully for exploring the local structure after identifying a feature of interest, perhaps via guided tour. Additionally, *spinifex* extends graphic outputs to *plotly* and *gganimation*. This work extends the functionality of and is compatible with *tourr*.

Keywords: grand tour, projection pursuit, manual tour, tourr, touring, high dimensional visualization, high dim vis, dimensionality reduction, visualization, statistical graphics, data-space.


## Introduction

<!--
What is a tour?
-->

<!--
Multivariate data analysis, and relationship with tour methods: PCA, LDA (t-SNE, local linear embeddings or MDS)
-->

<!--
What is a manual tour, and what is it useful for - DONE
-->

Both classical and contemporary visualizations of data are presented in ($d=2$) two dimensions, that of a computer monitor or in print. How is it that we come to view and share data that exists in $p > 3$ dimensions? In an appeal to brevity we shall ignore model and parameter summarization due to there shortcomings [@anscombe_graphs_1973; @matejka_same_2017]. Within the realm of data-space visualization we are left with projecting higher volumes and embedding them within lower dimensional spaces that we can visualize.

This is not a new phenomena, such linear projections have been in use for quite some time. [@pearson_liii._1901; @fisher_use_1936] and the myriad of single value decomposition (SVD) techniques from numerous disciplines use such embeddings. Previous application look at data in one (or few) static orientations, after some objective optimization. For instance in PCA, we reorient $p$-dimensions such that we have a reference to the ordered components that describe a descending amount of variation held within the data. Yet we still have $p$ components remaining to visualize. Where does the dimension reduction come in? From plotting only the first two or three and potentially another dimension tied to data point aesthetic. This is maximizes the amount of variation that can be display in an embedding, but regularly discards a large proportion of the variation held within the data.

More recently non-linear dimensionality techniques have become popular, such as t-distributed stochastic neighbor embedding (t-SNE) [@maaten_visualizing_2008], building off of Sammon mappings [@sammon_nonlinear_1969]. Such non-linear methods make for astounding distinction when in lower embeddings, but contain inherent shortcomings. Namely: that the non-linear transformations break inter-operability back to the original data-space, and that they can suffer from overfitting. If there is no inherent clustering with the data, it's possible that noise within the variables may become the prominent feature and be displayed erroneously as group clustering.

TODO: Clean up touring, , below

[@asimov_grand_1985; @buja_grand_1986] first suggested grand tours in which random walks in $p$-space can be interpolated and embedded in $d$ dimensions which are then viewed in sequence. Imagine. Consider, 


The broader scope of touring has some beneficial features, namely: touring keeps the original dimensionality in tact unlike tradition static linear-projections, and maintains inter-operability back to the original dimensions, a primary drawback of non-linear dimensionality reduction.

TODO: clean up above, talk about touring in general first

<!-- ## Terminology and demystifying projection: -->
<!-- TODO: clean up the terminology section -->
<!-- basis, data, n, p, d,  -->

<!-- Suppose that we have tri-variate data, $\textbf{X}_{[8,~3]}$, the corners points of a rectanguloid. We can describe the relative orientation by defining  -->

<!-- For every $p$-dimensional space can be described by the direction and magnitude of axes in a square matrix that we call a basis. Imgine 3 axes of an XYZ Caresian volume (*ie.* a basis $\in \mathbb{R}^p$). In matimatical form we would write this as a diagonal identity matrix of demension 3. -->

<!-- ```{r} -->
<!-- # b <- diag(3) -->
<!-- # xyz <- c("X", "Y", "Z") -->
<!-- # rownames(b) <- xyz -->
<!-- # colnames(b) <- xyz -->
<!-- # b -->
<!-- ``` -->

<!-- This basis has some nice properties that are mathimatically nice to preserve, namely, that each axis as is at a right angle to the other (*orthagonal*), and are unit *normal* (length or norm equal to one). If matrix meets both of these criteria we call it *orthonormal*. -->

<!-- ```{r} -->
<!-- # set.seed(15) -->
<!-- # library(plotly) -->
<!-- # library(processx) -->
<!-- # X = 9*runif(100) # X = 9*c(0,1,1,0,0,1,1,0) -->
<!-- # Y = 6*runif(100) # Y = 6*c(0,0,1,1,0,0,1,1) -->
<!-- # Z = 2*runif(100) # Z = 2*c(0,0,0,0,1,1,1,1) -->
<!-- #  -->
<!-- # p <- plot_ly(x=X, y=Y, z=Z, type="scatter3d", mode="markers") -->
<!-- # ###STATIC OUTPUT REQURIES ORCA SETUP, NOT TRANsFERABLE... -->
<!-- # # orca(p, "cube_demo.png") # but, needs orca setup,  -->
<!-- # # plotly_IMAGE(p, format = "png", out_file = "output.png") -->
<!-- ``` -->


The manual tour was described in @cook_manual_1997, and allows a user to rotate a variable into and out of a 2D projection of high-dimensional space. The primary purpose is to determine the sensitivity of structure visible in a projection to a variable. Manual touring can also be useful for exploring the local structure once a feature of interest has been identified, for example, by a guided tour [@hurley_analyzing_1990]. In guided tours an index of interest defined on the space of all projections, and the function is optimised. It derives from projection pursuit [@friedman_projection_1974], and the guided tour provides a visual interface to the optimisation. The manual tour can be used to help refine structure in the optimal projection, sharpening it by exploring the contributions of different variables, and simplifying by zeroing the coefficients of variables that don't contribute to the structure. 


## Manual tour algorithm

The algorithm contains several steps:

1. Choose a variable to explore, called the manip variable
2. Create a 3D manipulation space, where the variable to be explored has full contribution.
3. Generate the rotation which zero's the coefficient and also increases it to 1

### Notation

Given:
\begin{description}
  \item[$\textbf{X}_{[n,~p]}$] A data set containing $n$ observations of $p$ numeric variables. 
  \item[$\textbf{B}_{[p,~d]}$] An orthonormal \footnote{Where each variable is both: orthogonal, at right angles (dot product is 0) to the other variables, and unit vectors, a norm = 1} basis describing the current orientation projecting $p$ down to $d$ dimension.
\end{description}

\begin{align*}
  \textbf{X}_{[n,~p]} ~=
  \begin{bmatrix}
    X_{1,~1} & \dots  & X_{1,~p} \\
    X_{2,~1} & \dots  & X_{2,~p} \\
    \vdots   & \ddots & \vdots   \\
    X_{n,~1} & \dots  & X_{n,~p}
  \end{bmatrix}
\end{align*}

\begin{align*}
  \textbf{B}_{[p,~d]} ~=
  \begin{bmatrix}
    B_{1,~1} & \dots  & B_{1,~d} \\
    B_{2,~1} & \dots  & B_{2,~d} \\
    \vdots   & \ddots & \vdots   \\
    B_{p,~1} & \dots  & B_{p,~d}
  \end{bmatrix}
\end{align*}

For ease of computation we will be working mostly with the basis and not the data, once basis manipulation is done post multiply the data by the basis to get back to data-space.  

### Step 1 Choose variable of interest

Select a manipulation variable, $k$. Initialize a zero vector $e$, and set the $k$-th element set to 1. 

NOTE: Explanation has math, diagram and code

### Step 2 Create the manip space

NOTE: Explanation has math, diagram and code

Use the Gram-Schmidt process to orthonormalize the concatenation of the basis and $e$ yielding the manipulation space.

\begin{align*}
  \textbf{M}_{[p,~d+1]}
  &= Orthonormalize_{GS}( \textbf{B}_{[p,~d]}|\textbf{e}_{k~[p,~1]} ) \\
  &= Orthonormalize_{GS}
  \left(
    \begin{bmatrix}
      B_{1,~1} & \dots  & B_{1,~d} \\
      B_{2,~1} & \dots  & B_{2,~d} \\
      \vdots   & \ddots & \vdots   \\
      B_{k,~1} & \dots  & B_{k,~d} \\
      \vdots   & \ddots & \vdots   \\
      B_{p,~1} & \dots  & B_{p,~d}
    \end{bmatrix}
  ~|~
    \begin{bmatrix}
      0 \\
      0 \\
      \vdots \\
      1 \\
      \vdots \\
      0
    \end{bmatrix}
  \right)
\end{align*}

TODO: (below) clean up the language of phi, tie to manipulation var. Make sure to paint the picture for the  Z-direction of manip_sp. 

### Step 3 Generate rotation

NOTE: Explanation has math, diagram and code

Select a vector $\phi_i$, the angle of out-of plane rotation, orthogonal to the projection plane (relative to $\phi_1$, the transformation $\phi_i$ - $\phi_1$ proved to be helpful to discuss $\phi$ relative to the Z axis). 

**For ** $i$ **in 1 to n_slides:**
<!-- Don't fix the above, it's not going to happen... -->

For each $\phi_i$, post multiply the manipulation space by a rotation matrix,  producing as many basis-projections.

\begin{align*}
  \textbf{P}_{b[p,~d+1,~i]}
  &= \textbf{M}_{[p,~d+1]} ~*~ \textbf{R}_{[d+1,~d+1]} 
    ~~~~~~~~~~~~~~~~~~~\text{For the $d=2$ case:} \\
  &= \begin{bmatrix}
    M_{1,~1} & M_{1,~2} & M_{1,~3} \\
    M_{2,~1} & M_{2,~2} & M_{2,~3} \\
    \vdots   & \vdots   & \vdots   \\
    M_{p,~1} & M_{p,~2} & M_{p,~3}
  \end{bmatrix}_{[p,~d+1]}
    ~*~
  \begin{bmatrix}
    c_\theta^2 c_\phi s_\theta^2 &
    -c_\theta s_\theta (1 - c_\phi) &
    -c_\theta s_\phi \\
    -c_\theta s_\theta (1 - c_\phi) &
    s_\theta^2 c_\phi + c_\theta^2 &
    -s_\theta s_\phi \\
    c_\theta s_\phi &
    s_\theta s_\phi &
    c_\phi
  \end{bmatrix}_{[3,~3]}
\end{align*}


Where:
\begin{description}
  \item[$\theta$] is the angle that lies on the projection plane ($ie.$ on the XY plane)
  \item[$\phi$] is the angle orthogonal to the projection plane ($ie.$ in the Z direction)
  \item[$c_\theta$] is the cosine of $\theta$
  \item[$c_\phi$]   is the cosine of $\phi$
  \item[$s_\theta$] is the sine of   $\theta$
  \item[$s_\phi$]   is the sine of   $\phi$
\end{description}

To get back to data-space post multiply each projection basis by the data, for  the data projection.

\begin{align}
  \textbf{P}_{d[n,~d+1]}
    &= \textbf{X}_{[n,~p]} ~*~ \textbf{P}_{b[p,~d+1]} \\
    &= \begin{bmatrix}
      X_{1,~1} & \dots & X_{1,~p} \\
      X_{2,~1} & \dots & X_{2,~p} \\
      \vdots   & \vdots & \vdots  \\
      X_{n,~1} & \dots & X_{n,~p}
    \end{bmatrix}_{[n,~p]}
      ~*~
    \begin{bmatrix}
      P_{b:1,~1} & P_{b:1,~2} & P_{b:1,~3} \\
      P_{b:2,~1} & P_{b:2,~2} & P_{b:2,~3} \\
      \vdots     & \vdots     & \vdots     \\
      P_{b:p,~1} & P_{b:p,~2} & P_{b:p,~3}
    \end{bmatrix}_{b[p,~d+1]}
\end{align}

## Display projection sequence

<!--
- Lack of full interaction in R
- Animation packages
- Saving a sequence of projections, tourr, save_history
- Generating manual sequence off-line, replaying with animation package
-->

Plot the first two variables from each projection in sequence for an XY scatterplot. The remaining variable is sometimes linked to a data point aesthetic to produce depth cues used in conjunction with the XY scatterplot.


## Applcation

### Installation

spinifex is open source, and 

### Flea

Let's start off with the flea data set from the R package *tourr* [@wickham_tourr_2011], which performs different tours on the same data. The flea data contains 74 observations of flea beetles across 6 variables, physical measurements of the flea. Each individual belonged to one of three species being observed.

Let:
\begin{description}
  \item[$d=2$] For the sake of illustration we'll embed into two dimensions.
\end{description}

Given:
\begin{description}
  \item[$\bf{X}_{[74,~6]}$] Flea data, 74 observations by 6 variables, $\bf{X} \in \mathbb{R}^6, ~p=6$.
  \item[$p=6$]
\end{description}

Let's initialize a random orthonormal basis of dimensions [$p,~d$], which describes a random orientation projected from six down to two dimensions. Check how each of the dimensions is contributing the XY components with `view_basis()`

TODO: reference fig 1, keep in mind that the figure floats more than the table.
TODO: consider adding a `view_manip_sp()` function with phi and theata, etc.

```{r, fig.cap = "Random basis, flea data"}
library(spinifex)

flea_std <- rescale(flea[, 1:6])
rb <- basis_random(ncol(flea_std), 2)
view_basis(basis = rb, labels = colnames(flea_std)) 
```

Perform a manual tour on the random basis with `manual_tour()`, We'll arbitratrily choose the 4th variable, aede1, and let the default selection of `phi` go from it's start, to 0 radians, to pi/2 radians, and back to the start position. In turn the norm of manipualtion variable goes to 1 and then 0, before returning to it's inital position. 

```{r, fig.cap = "Manual tour, flea data", echo = TRUE, eval=FALSE}
mtour <- manual_tour(basis = rb, manip_var = 4)
mslides <- create_slides(tour = mtour, data = flea_std)

render_plotly(mslides)
```
 TODO: write follow up and segway into breast cancer; play_tour (holes), explore local area. 
 TODO: note that render_plotly didn't add the figure caption.

## Application

apply manual tours to Ursula fig 1 here:
NOTE: Examples from https://arxiv.org/abs/1806.09742

## Source

The code used to generate this chapter can be found on [GitHub] (https://github.com/nspyrison/Confirmation/blob/master/03-spinifex.Rmd).




## Discussion

Future work:

- other rotation mechanisms like Givens and Householder
- oblique should be defined.
- other display types, would work for 1D displays, but others would need redefinition difference dimensions
- shiny app, for choosing manip variable, providing different starting projections
